---
title: "Chatbot Ratings"
output: html_document
runtime: shiny
editor: visual
---

## Introduction

Chatbot Arena (lmarena.ai) is "an open-source platform for evaluating AI through human preference, developed by researchers at UC Berkeley [SkyLab](https://sky.cs.berkeley.edu/) and [LMSYS](https://lmsys.org/)." On the website <https://lmarena.ai/>, users can converse with two LLMs on any topic they choose. Users can then vote on which response they prefer (Model A or Model B) or can choose "Tie" or "Both bad" if they don't have a preference for either model. The LLMs are initially anonymous, but are revealed to the user after they vote.

## Data

The developers of this site provide data [here](https://huggingface.co/datasets/lmarena-ai/arena-human-preference-140k) on 140K conversations with pairwise human preferences. You can find this data in [this shared folder](https://yaleedu-my.sharepoint.com/:f:/g/personal/brian_macdonald_yale_edu/Eqo7cptmHAZAjNMv2F1dCj4BhZzs17e4hkfxbH8tMxLnwQ?e=Oo1DNN). There are 7 `.parquet` files. There is a README file in the GitHub repo that gives details about the data.

Each row is a matchup between two chat bots (e.g. Gemini 2.5 Pro vs Claude Sonnet 4), and the columns indicate a conversation ID, the names of the two models, who the user selected as the winner of the matchup, a common query that the two models were given, the category of the prompt (creative writing, math, code, etc.) and other information. See the README file for more details.

## Question

Which chatbot(s) appears to perform best in the eyes of the judges/users? In a matchup between any two models Model A and Model B, what are the outcome probabilities? How do these change if you focus on specific categories (e.g. writing, math, code, etc.) of prompts? Can you think of any reason the data may not be reliable, or any way this rating system could be "gamed"?

## Submission

Submit a PDF report with your code and analysis to Gradescope.

## Setup

```{r, eval=FALSE}
# read in data and bind rows
datalist <- vector(mode = "list", length = 7)

for(i in 0:6){
  data <- read_parquet(paste0("data/train-0000", i, "-of-00007.parquet")) %>% 
    as.data.frame() %>% 
    select(c(1:6, 11:14))
  datalist[[i+1]] <- data
  print(i)
}

df <- bind_rows(datalist)

flattened_data <- df %>% 
  unpack(cols = category_tag) %>%
  unpack(cols = c(creative_writing_v0.1, criteria_v0.1, if_v0.1, math_v0.1), 
         names_sep = "_")

saveRDS(flattened_data, file = "data.rds")
```

```{r}
# load packages
library(arrow)
library(dplyr)
library(tidyr)
library(stringr)
library(purrr)
library(ggplot2)
library(pubtheme)
library(tidyverse)
library(igraph)
library(ggraph)
library(tidygraph)
library(visNetwork)
library(scales)
library(viridisLite)
library(echarts4r)
library(BradleyTerry2)
library(shiny)
library(bslib)
library(tibble)
library(htmlwidgets)
library(DT)

# load data
data <- readRDS("data.rds")
head(data)
colnames(data)
```

## Categories

```{r}
data <- data %>%
  mutate(
    # Convert weird values to logicals / numeric flags
    cw_flag   = creative_writing_v0.1_creative_writing %in% c(TRUE, "TRUE", 1),
    cw_score  = suppressWarnings(as.numeric(creative_writing_v0.1_score)),
    
    math_flag = math_v0.1_math %in% c(TRUE, "TRUE", 1),
    
    code_flag = is_code %in% c(TRUE, "TRUE", 1),
    
    if_flag   = if_v0.1_if %in% c(TRUE, "TRUE", 1),
    if_score  = suppressWarnings(as.numeric(if_v0.1_score)),
    
    # Build category
    category = case_when(
      code_flag ~ "code",
      math_flag ~ "math",
      if_flag | if_score > 0 ~ "reasoning",
      cw_flag | cw_score > 0 ~ "creative_writing",
      TRUE ~ "other"
    )
  )
```

## Best Chatbot (Shiny App)

Includes overall leaderboard and option to filter by category and match-up model.

```{r}
all_categories <- sort(unique(data$category))

stopifnot("category" %in% names(data))

# Wilson CI for a proportion (used by Details table)
wilson_ci <- function(successes, n, z = 1.96) {
  ifelse(n > 0, {
    phat <- successes / n
    denom <- 1 + (z^2)/n
    center <- (phat + (z^2)/(2*n)) / denom
    half   <- (z * sqrt((phat*(1-phat) + (z^2)/(4*n))/n)) / denom
    tibble::tibble(lower = pmax(0, center - half), upper = pmin(1, center + half))
  }, tibble::tibble(lower = NA_real_, upper = NA_real_))
}

prof_all_models <- sort(unique(c(data$model_a, data$model_b)))
prof_all_categories <- sort(unique(data$category))

ui <- fluidPage(
  theme = bs_theme(bootswatch = "flatly"),
  sidebarLayout(
    sidebarPanel(
      class = "sticky-top",
      # View scope (Overall or a single category)
      selectInput(
        "view_mode", "View:",
        choices = c("Overall", all_categories),
        selected = "Overall"
      ),

      # Only show category multi-select when view_mode is Overall
      conditionalPanel(
        condition = "input.view_mode == 'Overall'",
        selectizeInput(
          "cat", "Filter categories (Overall):",
          choices = all_categories, selected = all_categories,
          multiple = TRUE, options = list(plugins = list("remove_button"))
        )
      ),

      # sort control
      selectInput(
        "sort_by", "Sort leaderboard by:",
        choices = c("Success rate", "Win rate"),
        selected = "Success rate"
      ),
      
      # Shared controls
      sliderInput("min_matches", "Min matches (for leaderboard rows)", min = 1, max = 500, value = 30, step = 1),
      
      # categories
      hr(),
      h4("Category Profile"),
      selectInput("prof_model", "Choose a model:", choices = prof_all_models, selected = prof_all_models[1]),
      selectizeInput(
        "prof_cats", "Categories:",
        choices = prof_all_categories, selected = prof_all_categories,
        multiple = TRUE, options = list(plugins = list("remove_button"))
      ),
      sliderInput("prof_min_n", "Min matches per category", min = 1, max = 200, value = 25, step = 1),
      checkboxInput("prof_use_success", "Use Success Rate (Win + 0.5×Tie)", value = TRUE),
      checkboxInput("prof_show_ci", "Show 95% CIs in table", value = TRUE),
      
      # head-to-head
      hr(),
      h4("Predicted vs Others"),
      selectInput("bt_model", "Choose a model:", choices = NULL)
    
    ),
    mainPanel(
      # ---- Leaderboard Card ----
      bslib::card(
        full_screen = TRUE, collapsible = TRUE, collapsed = FALSE,
        bslib::card_header(textOutput("main_title")),
        bslib::card_body(
          div(style = "font-weight:600; font-size:16px; margin-bottom:8px;",
              textOutput("subtitle")),
          echarts4rOutput("main_chart", height = "600px"),
          tags$hr(),
          div(style = "font-weight:600; margin:8px 0;", "Sortable table"),
          DTOutput("main_table")
        )
      ),
    
      # ---- Category Profile Card (combined) ----
      bslib::card(
        full_screen = TRUE, collapsible = TRUE, collapsed = FALSE,
        bslib::card_header(textOutput("prof_title_bar")),
        bslib::card_body(
          # Bar chart (model vs category average)
          echarts4rOutput("prof_bar", height = "520px"),
          tags$hr(),
    
          # Radar (normalized strengths)
          tags$div(style = "font-weight:600; margin-bottom:6px;",
                   "Strengths & Weaknesses (Radar; normalized within selected categories)"),
          echarts4rOutput("prof_radar", height = "420px"),
          tags$hr(),
    
          # Details table
          tags$div(style = "font-weight:600; margin:8px 0;", "Details by Category"),
          tableOutput("prof_detail_table")
        )
      ),
    
      # ---- Bradley–Terry (BT) Card ----
      bslib::card(
        full_screen = TRUE, collapsible = TRUE, collapsed = FALSE,
        bslib::card_header(textOutput("bt_title")),   # Title outside chart
        bslib::card_body(
          echarts4rOutput("bt_chart", height = "720px"),
          tags$br(),
          div(style = "font-weight:600; margin:8px 0;", "Predicted rates (pageable)"),
          DTOutput("bt_table")
        )
      )
    )
  )
)

server <- function(input, output, session){

  # ----- Data scoping (Overall vs one Category) ------------------------------
  selected_df <- reactive({
    if (identical(input$view_mode, "Overall")) {
      df <- data
      if (!is.null(input$cat) && length(input$cat) > 0) {
        df <- df %>% dplyr::filter(category %in% input$cat)
      }
      df
    } else {
      data %>% dplyr::filter(category == input$view_mode)
    }
  })
  
  # ---------- Category Profile module (independent of leaderboard/BT) ----------
  
  # Filter to chosen categories for the profile block
  prof_filtered <- reactive({
    req(input$prof_cats)
    df <- data %>% dplyr::filter(category %in% input$prof_cats)
    validate(need(nrow(df) > 0, "No rows in selected categories."))
    df
  })
  
  # Long form per (model, category) with ties only (both_bad NOT counted as tie here)
  prof_df2 <- reactive({
    prof_filtered() %>%
      dplyr::mutate(
        a_win = as.integer(winner == "model_a"),
        b_win = as.integer(winner == "model_b"),
        a_tie = as.integer(winner == "tie"),
        b_tie = a_tie
      ) %>%
      tidyr::pivot_longer(c(model_a, model_b), names_to = "side", values_to = "model") %>%
      dplyr::mutate(
        win = dplyr::if_else(side == "model_a", a_win, b_win),
        tie = dplyr::if_else(side == "model_a", a_tie, b_tie)
      )
  })
  
  # Per-(model, category) summary
  prof_model_cat_summary <- reactive({
    prof_df2() %>%
      dplyr::group_by(model, category) %>%
      dplyr::summarise(
        wins = sum(win),
        ties = sum(tie),
        matches = dplyr::n(),
        win_rate = wins / matches,
        tie_rate = ties / matches,
        success_rate = (wins + 0.5 * ties) / matches,
        .groups = "drop"
      )
  })
  
  # Selected model profile + category benchmark (weighted by per-model matches)
  prof_sel_profile <- reactive({
    req(input$prof_model)
    metric <- if (isTRUE(input$prof_use_success)) "success_rate" else "win_rate"
    mcs <- prof_model_cat_summary()
  
    me <- mcs %>%
      dplyr::filter(model == input$prof_model) %>%
      dplyr::rename(model_matches = matches)
  
    bench <- mcs %>%
      dplyr::group_by(category) %>%
      dplyr::summarise(
        bench = stats::weighted.mean(.data[[metric]], w = .data$matches),
        total_matches = sum(.data$matches),
        .groups = "drop"
      )
  
    me %>%
      dplyr::inner_join(bench, by = "category") %>%
      dplyr::mutate(
        metric = .data[[metric]],
        lift = metric - bench
      ) %>%
      dplyr::filter(model_matches >= input$prof_min_n) %>%
      dplyr::arrange(dplyr::desc(metric))
  })
  
  # Title
  output$prof_title_bar <- renderText({
    paste0(
      "Category Profile for ", input$prof_model, " — ",
      if (isTRUE(input$prof_use_success)) "Success Rate (Win + 0.5×Tie)" else "Win Rate",
      "  |  Min matches: ", input$prof_min_n
    )
  })
  
  # BAR: model vs category average (with lift tooltip)
  output$prof_bar <- renderEcharts4r({
    prof <- prof_sel_profile()
  
    prof <- prof %>%
      dplyr::filter(is.finite(metric), is.finite(bench)) %>%
      dplyr::mutate(
        metric = ifelse(is.na(metric), 0, metric),
        bench  = ifelse(is.na(bench),  0, bench)
      )
  
    validate(need(nrow(prof) > 0, "No categories meet the minimum matches threshold."))
  
    chart_df <- prof %>%
      dplyr::arrange(dplyr::desc(metric)) %>%
      dplyr::mutate(
        cat_lab = sprintf("%s  (n=%d)", category, model_matches),
        my_val  = round(metric, 6),
        bench_r = round(bench,  6)
      )
  
    chart_df |>
      e_charts(cat_lab) |>
      e_bar(
        my_val, name = input$prof_model,
        label = list(
          show = TRUE, position = "insideRight",
          color = "#fff", fontSize = 10,
          formatter = htmlwidgets::JS(
            "function(p){
               if(p.value == null || !isFinite(p.value)) return '';
               return (p.value*100).toFixed(0) + '%';
             }")
        )
      ) |>
      e_line(
        bench_r, name = "Category Avg", y_index = 0,
        symbol = "circle", symbolSize = 6
      ) |>
      e_tooltip(
        trigger = "axis",
        formatter = htmlwidgets::JS(
          "function (params) {
             const bar  = params.find(p => p.seriesName !== 'Category Avg');
             const line = params.find(p => p.seriesName === 'Category Avg');
             const vBar  = (bar  && isFinite(bar.value))  ? bar.value  : 0;
             const vLine = (line && isFinite(line.value)) ? line.value : 0;
             const lift  = (vBar - vLine)*100;
             return params[0].axisValue + '<br/>' +
                    bar.seriesName + ': ' + (vBar*100).toFixed(1) + '%<br/>' +
                    'Category Avg: ' + (vLine*100).toFixed(1) + '%<br/>' +
                    'Lift: ' + lift.toFixed(1) + ' pts';
           }")
      ) |>
      e_flip_coords() |>
      e_y_axis(inverse = TRUE) |>
      e_title("Model vs Category Average") |>
      e_x_axis(
        max = 1,
        axisLabel = list(formatter = htmlwidgets::JS(
          "function(x){ return (x*100).toFixed(0)+'%'; }"))
      ) |>
      e_axis_labels(
        y = "Category (n = matches)",
        x = if (isTRUE(input$prof_use_success)) "Success Rate" else "Win Rate"
      ) |>
      e_grid(left = 170) |>
      e_legend(right = 10, top = 30) |>
      e_datazoom(y_index = 0, type = "slider") |>
      e_theme("infographic") |>
      e_toolbox_feature("saveAsImage")
  })
  
  # RADAR: normalized strengths (0–1 within selected categories)
  output$prof_radar <- renderEcharts4r({
    mcs <- prof_model_cat_summary() %>% dplyr::filter(matches >= input$prof_min_n)
    metric <- if (isTRUE(input$prof_use_success)) "success_rate" else "win_rate"
  
    cat_minmax <- mcs %>%
      dplyr::group_by(category) %>%
      dplyr::summarise(
        mn = suppressWarnings(min(.data[[metric]], na.rm = TRUE)),
        mx = suppressWarnings(max(.data[[metric]], na.rm = TRUE)),
        .groups = "drop"
      )
  
    me <- mcs %>%
      dplyr::filter(model == input$prof_model) %>%
      dplyr::inner_join(cat_minmax, by = "category") %>%
      dplyr::mutate(
        val = dplyr::case_when(
          is.finite(mx) & is.finite(mn) & mx > mn ~ ( .data[[metric]] - mn ) / (mx - mn),
          TRUE ~ 0.5
        )
      ) %>%
      dplyr::arrange(category) %>%
      dplyr::filter(is.finite(val))
  
    validate(need(nrow(me) >= 3, "Not enough categories with sufficient matches for radar."))
  
    df_wide <- me %>%
      dplyr::select(category, val) %>%
      tidyr::pivot_wider(names_from = category, values_from = val) %>%
      dplyr::mutate(across(dplyr::everything(), ~ ifelse(is.na(.x), 0.5, .x))) %>%
      dplyr::mutate(model = input$prof_model)
  
    cat_cols <- setdiff(names(df_wide), "model")
    validate(need(length(cat_cols) >= 3, "Radar needs ≥ 3 categories."))
  
    indicator_list <- lapply(cat_cols, function(nm) list(name = nm, max = 1))
  
    df_wide |>
      e_charts(model) |>
      e_radar(!!!rlang::syms(cat_cols), name = input$prof_model) |>
      e_radar_opts(indicator = indicator_list) |>
      e_title("Normalized Strengths by Category (0–1 within chosen categories)") |>
      e_tooltip(trigger = "item") |>
      e_theme("infographic")
  })
  
  # DETAILS table (with optional 95% Wilson CI)
  output$prof_detail_table <- renderTable({
    prof <- prof_sel_profile()
    metric <- if (isTRUE(input$prof_use_success)) "success_rate" else "win_rate"
  
    if (isTRUE(input$prof_use_success)) {
      ci <- wilson_ci(successes = prof$wins + 0.5 * prof$ties, n = prof$model_matches)
    } else {
      ci <- wilson_ci(successes = prof$wins, n = prof$model_matches)
    }
  
    tibble::tibble(
      Category = prof$category,
      `Matches (model)` = prof$model_matches,
      Wins = prof$wins,
      Ties = prof$ties,
      Rate = scales::percent(prof$metric, accuracy = 0.1),
      `95% CI` = if (isTRUE(input$prof_show_ci))
        paste0(scales::percent(ci$lower, 0.1), " – ", scales::percent(ci$upper, 0.1))
        else NA_character_,
      `Category Avg` = scales::percent(prof$bench, 0.1),
      `Lift (pts)` = round(100 * prof$lift, 1),
      `Matches (category total)` = prof$total_matches
    )
  },
  striped = TRUE,
  bordered = FALSE,
  spacing = "m",
  align = "lrrrccrrr"
  )
  
  # BT head-to-head
  bt_scoped_data <- reactive({
    # Fit on ALL scoped matches; no global min-matches here.
    selected_df()
  })
  
  bt_scope_label <- reactive({
    if (identical(input$view_mode, "Overall")) {
      cats <- if (is.null(input$cat) || !length(input$cat)) "All categories" else paste(input$cat, collapse = ", ")
      paste0("Overall (", cats, ")")
    } else {
      paste0("Category: ", input$view_mode)
    }
  })
  
  bt_title_text <- reactive({
    paste0(
      "Predicted probability of winning against ", input$bt_model,
      " (min matches = ", input$min_matches, ") | ",
      bt_scope_label()
    )
  })
  
  output$bt_title <- renderText(bt_title_text())
  
  # Order models by simple empirical success to make the picker friendly
  bt_model_order <- reactive({
    df <- selected_df()
    appear <- df %>%
      dplyr::select(model_a, model_b) %>%
      tidyr::pivot_longer(everything(), values_to = "model") %>%
      dplyr::count(model, name = "appearances")
    wins <- df %>%
      dplyr::filter(winner %in% c("model_a","model_b")) %>%
      dplyr::mutate(winner_model = dplyr::if_else(winner == "model_a", model_a, model_b)) %>%
      dplyr::count(winner_model, name = "wins") %>%
      dplyr::rename(model = winner_model)
    appear %>%
      dplyr::left_join(wins, by = "model") %>%
      dplyr::mutate(wins = tidyr::replace_na(wins, 0),
                    success_rate = wins / appearances) %>%
      dplyr::arrange(dplyr::desc(success_rate)) %>%
      dplyr::pull(model)
  })
  
  observeEvent(bt_model_order(), {
    choices <- bt_model_order()
    # If no model passes the filter, keep it empty; otherwise select first valid
    sel <- if (length(choices) && input$bt_model %in% choices) input$bt_model else dplyr::first(choices)
    updateSelectInput(session, "bt_model", choices = choices, selected = sel)
  }, ignoreInit = FALSE)
  
  # Fit Bradley–Terry model on decisive matches in the current scope
  bt_bt_fit <- reactive({
    pair_outcomes <- bt_pair_outcomes()
    validate(need(nrow(pair_outcomes) > 0, "No pairwise outcomes available in the selected filter."))
  
    # --- Keep only the connected component containing the selected model ----
    edges <- pair_outcomes %>% dplyr::transmute(from = A, to = B)
    g <- igraph::graph_from_data_frame(edges, directed = FALSE)
  
    keep_names <- {
      comps <- igraph::components(g)
      if (isTRUE(input$bt_model %in% igraph::V(g)$name)) {
        vid <- which(igraph::V(g)$name == input$bt_model)
        comp_id <- comps$membership[vid]
        names(which(comps$membership == comp_id))
      } else {
        names(which(comps$membership == which.max(comps$csize)))
      }
    }
  
    pair_cc <- pair_outcomes %>%
      dplyr::filter(A %in% keep_names, B %in% keep_names)
    validate(need(nrow(pair_cc) > 0, "Selected model is isolated under current filters."))
  
    players <- sort(unique(c(pair_cc$A, pair_cc$B)))
  
    # Smoothing to reduce separation risk (don’t invent ties)
    eps <- 0.5
    A_win_s <- pair_cc$A_win + eps
    B_win_s <- pair_cc$B_win + eps
    tie_s   <- pair_cc$tie
  
    has_ties <- sum(tie_s, na.rm = TRUE) > 0
    y <- if (has_ties) cbind(A_win_s, B_win_s, tie_s) else cbind(A_win_s, B_win_s)
  
    fit <- tryCatch(
      BradleyTerry2::BTm(
        outcome = y,
        player1 = factor(pair_cc$A, levels = players),
        player2 = factor(pair_cc$B, levels = players),
        id = "player",
        ties = has_ties
      ),
      error = function(e) NULL
    )
  
    # If still NULL, return an object that signals "use empirical fallback"
    if (is.null(fit)) {
      return(list(
        fit = NULL,
        has_ties = has_ties,
        players = players,
        mode = "empirical",
        pair_cc = pair_cc
      ))
    }
  
    list(
      fit = fit,
      has_ties = has_ties,
      players = players,
      mode = "model",
      pair_cc = pair_cc
    )
  })
  
  bt_abilities <- reactive({
    fit <- bt_bt_fit()
    ab <- BradleyTerry2::BTabilities(fit) %>% as.data.frame() %>%
      tibble::rownames_to_column("model") %>% dplyr::select(model, ability)
    setNames(ab$ability, ab$model)
  })
  
  bt_pair_counts <- reactive({
    selected_df() %>%
      dplyr::transmute(A = pmin(model_a, model_b),
                       B = pmax(model_a, model_b)) %>%
      dplyr::filter(A != B) %>%
      dplyr::summarise(matches = dplyr::n(), .by = c(A, B))
  })
  
  bt_pair_outcomes <- reactive({
    selected_df() %>%
      dplyr::transmute(
        A = pmin(model_a, model_b),
        B = pmax(model_a, model_b),
        res = dplyr::case_when(
          winner %in% c("tie", "both_bad") ~ "tie",
          winner == "model_a" & model_a <= model_b ~ "A_win",
          winner == "model_b" & model_a  >  model_b ~ "A_win",
          winner == "model_b" & model_a <= model_b ~ "B_win",
          winner == "model_a" & model_a  >  model_b ~ "B_win",
          TRUE ~ NA_character_
        )
      ) %>%
      dplyr::filter(!is.na(res), A != B) %>%
      dplyr::count(A, B, res) %>%
      tidyr::pivot_wider(
        names_from = res, values_from = n, values_fill = 0
      )
  })
  
  # Build vs-opponents table from abilities + tie rates
  bt_make_vs_btd <- function(fit_obj, sel, model_choices, pair_counts) {
    # Ensure we have something to compare to
    opps_all <- setdiff(model_choices, sel)
    if (!length(opps_all)) {
      return(tibble::tibble(
        opponent = character(), matches_vs_selected = integer(),
        p_sel_vs_row = numeric(), p_row_vs_sel = numeric(),
        p_tie_or_both_bad = numeric()
      ))
    }
  
    players <- fit_obj$players
    opps <- intersect(opps_all, players)
    if (!length(opps)) {
      return(tibble::tibble(
        opponent = character(), matches_vs_selected = integer(),
        p_sel_vs_row = numeric(), p_row_vs_sel = numeric(),
        p_tie_or_both_bad = numeric()
      ))
    }
  
    # Join total match counts for sel vs opponent (for per-pair threshold & display)
    base <- tibble::tibble(
      opponent = opps,
      A = pmin(sel, opps), B = pmax(sel, opps)
    ) %>%
      dplyr::left_join(pair_counts, by = c("A","B")) %>%
      dplyr::mutate(matches_vs_selected = tidyr::replace_na(matches, 0L)) %>%
      dplyr::select(opponent, A, B, matches_vs_selected)
  
    if (identical(fit_obj$mode, "model") && !is.null(fit_obj$fit)) {
      # Use BT/BTD predicted probabilities
      nd <- tibble::tibble(
        player1 = factor(rep(sel, length(opps)), levels = players),
        player2 = factor(opps, levels = players)
      )
      pr <- as.data.frame(predict(fit_obj$fit, newdata = nd, type = "response"))
      colnames(pr) <- tolower(colnames(pr))
      p1 <- pr[["player1"]]
      p2 <- pr[["player2"]]
      pt <- if ("tie" %in% colnames(pr)) pr[["tie"]] else rep(0, length(p1))
  
      return(base %>%
        dplyr::mutate(
          p_sel_vs_row = p1,
          p_row_vs_sel = p2,
          p_tie_or_both_bad = pt
        ) %>%
        dplyr::select(opponent, matches_vs_selected, p_sel_vs_row, p_row_vs_sel, p_tie_or_both_bad) %>%
        dplyr::arrange(dplyr::desc(p_row_vs_sel)))
    }
  
    # --- Empirical Dirichlet(1/2,1/2,1/2) fallback (guaranteed finite) ----
    # Get the per-pair outcome counts inside the fitted component
    pair_cc <- fit_obj$pair_cc %>%
      dplyr::select(A, B, A_win, B_win, tie)
  
    alpha <- 0.5  # Jeffreys prior
    emp <- base %>%
      dplyr::left_join(pair_cc, by = c("A","B")) %>%
      dplyr::mutate(
        A_win = tidyr::replace_na(A_win, 0L),
        B_win = tidyr::replace_na(B_win, 0L),
        tie   = tidyr::replace_na(tie,   0L),
        total = A_win + B_win + tie,
        pA = (A_win + alpha) / (total + 3 * alpha),
        pB = (B_win + alpha) / (total + 3 * alpha),
        pT = (tie   + alpha) / (total + 3 * alpha),
        # map to selected vs row regardless of lexical order
        p_sel_vs_row = dplyr::if_else(sel <= opponent, pA, pB),
        p_row_vs_sel = dplyr::if_else(sel <= opponent, pB, pA),
        p_tie_or_both_bad = pT
      ) %>%
      dplyr::select(opponent, matches_vs_selected, p_sel_vs_row, p_row_vs_sel, p_tie_or_both_bad) %>%
      dplyr::arrange(dplyr::desc(p_row_vs_sel))
  
    emp
  }

  # chart
  output$bt_chart <- renderEcharts4r({
    fit_obj  <- bt_bt_fit()
    pair_cnt <- bt_pair_counts()
    choices  <- bt_model_order()
    validate(need(length(choices) > 0, "No models available for the selected filter."))
  
    df_all <- bt_make_vs_btd(fit_obj, input$bt_model, choices, pair_cnt)
  
    # Per-pair threshold + clean values
    df <- df_all %>%
      dplyr::filter(matches_vs_selected >= input$min_matches) %>%
      tidyr::replace_na(list(p_row_vs_sel = 0, p_tie_or_both_bad = 0)) %>%
      dplyr::mutate(
        p_row_vs_sel      = pmin(pmax(p_row_vs_sel, 0), 1),
        p_tie_or_both_bad = pmin(pmax(p_tie_or_both_bad, 0), 1),
        rank  = dplyr::row_number(),
        label = sprintf("#%d  %s", rank, opponent)
      )
  
    validate(need(nrow(df) > 0, "No head-to-head predictions meet the min-matches threshold."))
  
    # Default view: Top-10 via dataZoom
    start_lab <- df$label[1]
    end_lab   <- df$label[min(10, nrow(df))]
  
    df |>
      e_charts(label) |>
      # STACKED: Row beats Selected + Tie/both_bad
      e_bar(p_row_vs_sel,
            name  = "P(Row beats Selected)",
            stack = "prob",
            label = list(show = TRUE, position = "insideRight", fontSize = 10)) |>
      e_bar(p_tie_or_both_bad,
            name  = "P(Tie / both_bad)",
            stack = "prob",
            label = list(show = TRUE, position = "right", fontSize = 10)) |>
      e_flip_coords() |>
      e_y_axis(inverse = TRUE) |>
      e_legend(show = TRUE, right = 10, top = 40) |>
      e_title(paste0("Predicted Probabilities vs ", input$bt_model,
                     " — Row Win + Tie/both_bad (stacked)")) |>
      e_datazoom(y_index = 0, type = "slider", startValue = start_lab, endValue = end_lab) |>
      e_datazoom(y_index = 0, type = "inside", startValue = start_lab, endValue = end_lab) |>
      e_x_axis(
        max = 1,
        axisLabel = list(formatter = htmlwidgets::JS("function(x){return (x*100).toFixed(0)+'%';}"))
      ) |>
      e_axis_labels(
        y = "Row Model (ranked by P(row beats selected))",
        x = "Predicted Probability"
      ) |>
      e_grid(left = 240) |>
      e_tooltip(show = FALSE) |>
      e_theme("infographic") |>
      e_toolbox_feature("saveAsImage")
  })

  # Pageable table of rates
  output$bt_table <- DT::renderDT({
    fit      <- bt_bt_fit()
    pair_cnt <- bt_pair_counts()
    choices  <- bt_model_order()
    validate(need(length(choices) > 0, "No models available for the selected filter."))
  
    tbl <- bt_make_vs_btd(fit, input$bt_model, choices, pair_cnt) %>%
      dplyr::filter(matches_vs_selected >= input$min_matches)
    validate(need(nrow(tbl) > 0, "No head-to-head rows meet the min-matches threshold."))
  
    DT::datatable(
      tbl,
      rownames = FALSE,
      extensions = "Buttons",
      options = list(
        pageLength = 10,                  # default Top-10 via paging
        lengthMenu = c(5, 10, 25, 50, 100),
        order = list(list(3, "desc")),    # 0-based: p_row_vs_sel
        dom = "Bfrtip",
        buttons = c("copy", "csv", "excel")
      )
    ) %>%
      DT::formatPercentage(
        columns = c("p_sel_vs_row", "p_row_vs_sel", "p_tie_or_both_bad"),
        digits = 1
      )
  })

  # ----- Leaderboard (per model) ---------------------------------------------
  make_df2 <- function(df) {
    df %>%
      mutate(
        a_win = as.integer(winner == "model_a"),
        b_win = as.integer(winner == "model_b"),
        a_tie = as.integer(winner == "tie"),   # <- ONLY "tie" counts as tie
        b_tie = a_tie                          #    "both_bad" becomes a loss for both
      ) %>%
      pivot_longer(c(model_a, model_b), names_to = "side", values_to = "model") %>%
      mutate(
        win = if_else(side == "model_a", a_win, b_win),
        tie = if_else(side == "model_a", a_tie, b_tie)
      )
  }

  leaderboard <- reactive({
    df2 <- make_df2(selected_df())
    sort_col <- if (identical(input$sort_by, "Win rate")) "win_rate" else "success_rate"
  
    df2 %>%
      group_by(model) %>%
      summarise(
        wins = sum(win),
        ties = sum(tie),
        matches = n(),
        win_rate = wins / matches,
        tie_rate = ties / matches,
        success_rate = (wins + 0.5 * ties) / matches,
        .groups = "drop"
      ) %>%
      filter(matches >= input$min_matches) %>%
      arrange(dplyr::desc(.data[[sort_col]])) %>%
      mutate(
        rank     = row_number(),
        label    = sprintf("#%d  %s", rank, model),
        win_comp = round(win_rate, 3),
        tie_half = round(0.5 * tie_rate, 3)
      )
  })

  # ----- Titles ---------------------------------------------------------------
  output$main_title <- renderText({
    scope <- if (identical(input$view_mode, "Overall")) {
      cats <- if (is.null(input$cat) || !length(input$cat)) "All categories"
              else paste(input$cat, collapse = ", ")
      paste0("Overall (", cats, ")")
    } else paste0("Category: ", input$view_mode)
  
    paste0("Leaderboard — Win + 0.5×Tie (min matches = ", input$min_matches,
           ") | ", scope, " | Sorted by: ", input$sort_by)
  })

  # ----- Chart (Leaderboard only) --------------------------------------------
  output$main_chart <- renderEcharts4r({
    lb <- leaderboard()
    validate(need(nrow(lb) > 0, "No models meet the minimum matches threshold."))

    # dynamic title / x-axis label based on sort choice
    metric_title <- if (identical(input$sort_by, "Win rate")) "Win Rate" else "Success Rate (Win + 0.5×Tie)"
    x_label      <- if (identical(input$sort_by, "Win rate")) "Win Rate" else "Success Rate"

    # Default zoom to top 10
    start_lab <- lb$label[1]
    end_lab   <- lb$label[min(10, nrow(lb))]

    lb |>
      e_charts(label) |>
      e_bar(win_comp, name = "Win Rate", stack = "rate",
            label = list(show = TRUE, position = "insideRight", color = "#fff", fontSize = 10)) |>
      e_bar(tie_half, name = "0.5*Tie Rate", stack = "rate",
            label = list(show = TRUE, position = "right", color = "#000", fontSize = 10)) |>
      e_flip_coords() |>
      e_y_axis(inverse = TRUE) |>
      e_legend(orient = "vertical", right = 10, top = 40) |>
      e_title(paste0("Chatbots Ranked by ", metric_title)) |>
      e_datazoom(y_index = 0, type = "slider", startValue = start_lab, endValue = end_lab) |>
      e_datazoom(y_index = 0, type = "inside", startValue = start_lab, endValue = end_lab) |>
      e_x_axis(max = 1,
               axisLabel = list(formatter = htmlwidgets::JS("function(x){return (x*100).toFixed(0)+'%';}"))) |>
      e_axis_labels(y = "Model (ranked)", x = x_label) |>
      e_grid(left = 170) |>
      e_tooltip(show = FALSE) |>
      e_theme("infographic") |>
      e_toolbox_feature("saveAsImage")
  })

  # ----- Table (Leaderboard only) --------------------------------------------
  output$main_table <- renderDT({
    lb <- leaderboard()
    validate(need(nrow(lb) > 0, "No models meet the minimum matches threshold."))
    tbl <- lb %>% select(model, wins, ties, matches, win_rate, tie_rate, success_rate)

    # 0-based column index for DT ordering
    order_col <- if (identical(input$sort_by, "Win rate")) 4L else 6L

    datatable(
      tbl,
      rownames = FALSE,
      extensions = "Buttons",
      options = list(
        pageLength = 10,
        lengthMenu = c(5, 10, 25, 50, 100),
        order = list(list(order_col, "desc")),
        dom = "Bfrtip",
        buttons = c("copy", "csv", "excel")
      )
    ) %>%
      formatPercentage(columns = c("win_rate", "tie_rate", "success_rate"), digits = 1)
  })
}

shinyApp(ui, server)
```

```{r}
# 1) order by success_rate (highest → lowest)
model_order <- model_summary %>% arrange(desc(success_rate)) %>% pull(model)

# keep only models present in the BT abilities table
model_order <- intersect(model_order, abilities$model)

# 2) ability lookup
ab_map <- setNames(abilities$ability, abilities$model)

# 3) full grid of pairs with BT prob P(row beats col)
grid <- expand.grid(
  row_model = model_order,
  col_model = model_order,
  stringsAsFactors = FALSE
) %>%
  mutate(
    prob = ifelse(
      row_model == col_model, NA_real_,
      plogis(ab_map[row_model] - ab_map[col_model])  # exp(a)/(exp(a)+exp(b))
    )
  )

# factors to get the desired ordering in the plot:
grid$row_model <- factor(grid$row_model, levels = rev(model_order))  # rows: top = best
grid$col_model <- factor(grid$col_model, levels = model_order)       # cols: left = best

library(plotly)
library(scales)

grid <- grid %>% mutate(
  hover = sprintf("Row: %s<br>Col: %s<br>P(Row beats Col): %s",
                  row_model, col_model,
                  ifelse(is.na(prob), "—", percent(prob, 0.1)))
)

p <- ggplot(grid, aes(x = col_model, y = row_model, fill = prob, text = hover)) +
  geom_tile(color = "white", linewidth = 0.2, na.rm = FALSE) +
  scale_fill_viridis_c(option = "D", limits = c(0,1),
                       labels = percent_format(accuracy = 1),
                       name = "P(Row beats Col)") +
  coord_equal() +
  labs(title = "Head-to-Head Win Probabilities (Bradley–Terry)",
       x = "Column model", y = "Row model") +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(size = 4, angle = 45, hjust = 1, vjust = 1),
        axis.text.y = element_text(size = 4))

ggplotly(p, tooltip = "text")

```

```{r}
# edges and nodes
pairs <- data %>% 
  count(model_a, model_b, name = "matches") %>%
  mutate(
    edge_width = log1p(matches),   # damp outliers for thickness
    match_bin = cut(                # edge COLOR uses raw-match bins (for legend)
      matches,
      breaks = c(1,5,10,25,50,100,Inf),
      labels = c("1–5","6–10","11–25","26–50","51–100","100+"),
      right = TRUE, include.lowest = TRUE
    )
  )

nodes <- model_summary %>% 
  select(model, success_rate)

g <- graph_from_data_frame(d = pairs, vertices = nodes, directed = FALSE)

# ---- ADDED: total matches per model (sum of incident edge matches) ----------
totals <- pairs %>%
  select(model_a, model_b, matches) %>%
  pivot_longer(c(model_a, model_b), names_to = "side", values_to = "model") %>%
  group_by(model) %>%
  summarise(total_matches = sum(matches, na.rm = TRUE), .groups = "drop")

# success rate bins for legend + node popups
nodes2 <- model_summary %>%
  select(model, success_rate) %>%
  rename(id = model) %>%
  left_join(totals, by = c("id" = "model")) %>%     # ADDED
  mutate(
    success_bin = cut(
      success_rate,
      breaks = seq(min(success_rate, na.rm = TRUE),  # ADDED: na.rm for safety
                   max(success_rate, na.rm = TRUE),
                   length.out = 7),
      include.lowest = TRUE
    ),
    color = viridisLite::viridis(6)[as.integer(success_bin)],
    label = id,
    font.color = color,
    # ADDED: tooltip content
    title = paste0(
      "<b>", id, "</b><br>",
      "Success rate: ", percent(success_rate, accuracy = 0.1), "<br>",
      "Total matches: ", ifelse(is.na(total_matches), 0, total_matches)
    )
  )

# --- NEW: empirical head-to-head outcome probabilities per unordered pair ----
edge_probs <- data %>%
  transmute(
    A = pmin(model_a, model_b),
    B = pmax(model_a, model_b),
    res = case_when(
      winner %in% c("tie","both_bad") ~ "tie",
      winner == "model_a" & model_a <= model_b ~ "A_win",
      winner == "model_b" & model_a  > model_b ~ "A_win",
      winner == "model_b" & model_a <= model_b ~ "B_win",
      winner == "model_a" & model_a  > model_b ~ "B_win",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(res), A != B) %>%
  count(A, B, res) %>%
  tidyr::pivot_wider(
    names_from = res, values_from = n, values_fill = 0
  ) %>%
  mutate(
    total = A_win + B_win + tie,
    pA = ifelse(total > 0, A_win / total, NA_real_),  # P(A beats B)
    pB = ifelse(total > 0, B_win / total, NA_real_),  # P(B beats A)
    pT = ifelse(total > 0, tie   / total, NA_real_)   # tie rate
  )

# --- NEW: join probabilities onto your existing pairs in a direction-aware way
pairs_with_probs <- pairs %>%
  mutate(A = pmin(model_a, model_b),
         B = pmax(model_a, model_b)) %>%
  left_join(edge_probs, by = c("A","B")) %>%
  mutate(
    from = model_a, to = model_b,
    p_from = if_else(model_a <= model_b, pA, pB),
    p_to   = if_else(model_a <= model_b, pB, pA),
    tie_rate = pT
  )

# darker greys, first one darker manually (keep your palette)
edge_palette <- RColorBrewer::brewer.pal(6, "Greys")
edge_palette[1] <- "#6f6f6f"

vn_edges2 <- pairs_with_probs %>%
  transmute(
    from, to,
    width = pmax(1, round(log1p(matches))),
    color = edge_palette[as.integer(match_bin)],
    # Edge label now includes matchups + probabilities (rounded to 2dp)
    title = sprintf(
      "<b>%s ↔ %s</b><br>Matches: %d<br>%s win prob: %.2f%%<br>%s win prob: %.2f%%%s",
      from, to, matches,
      from, 100 * p_from,
      to,   100 * p_to,
      ifelse(is.na(tie_rate), "", sprintf("<br>Tie rate: %.2f%%", 100 * tie_rate))
    ),
    smooth = FALSE
  )

visNetwork(nodes2, vn_edges2, width = "100%", height = "800px") %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visPhysics(enabled = FALSE) %>%
  visInteraction(dragView = TRUE, zoomView = TRUE) %>%
  visOptions(
    nodesIdSelection = TRUE,                         # ADDED (search box)
    highlightNearest = list(enabled = TRUE, degree = 1, hover = TRUE)  # ADDED (nice-to-have)
  ) %>%
  visLegend(
    addEdges = data.frame(
      label = levels(pairs$match_bin),
      color = edge_palette
    ),
    addNodes = data.frame(
      label = levels(nodes2$success_bin),
      color = viridisLite::viridis(6)
    ),
    useGroups = FALSE,
    width = 0.25
  )
```

## Limitations

```{r}
# Count matches per pair and per model — sparsity:
df2 %>% count(model_a, model_b) %>% arrange(n)  # look for tiny n
model_summary %>% arrange(matches)  # models with very few matches

# votes per session
session_votes <- flat_data %>%
  count(evaluation_session_id)

ggplot(session_votes, aes(n)) +
  geom_histogram(bins = 40) +
  pub_theme() +
  labs(title = "Votes per Session",
       x = "Votes in Session", y = "Count of Sessions")

# Distribution of judges / sessions: are a small set of users contributing many votes? If you have evaluation_session_id or timestamp, check duplicates:
df2 %>% count(evaluation_session_id) %>% top_n(20, n)

# Order effect (evaluation_order): does the first-shown model get more votes?
df2 %>% group_by(evaluation_order) %>%
  summarise(win_rate_first = mean(if_else(evaluation_order == "A", a_win, b_win)))

# Time trends: models have versions; check timestamps — a model may improve over time.
library(lubridate)
df2 %>% mutate(ts = ymd_hms(timestamp)) %>% arrange(ts) %>% 
  NULL

model_ts <- flat_data %>%
  filter(model_a == "Claude 3.5 Sonnet") %>%   # example — you can parametrize this
  mutate(date = as.Date(timestamp)) %>%
  group_by(date) %>%
  summarise(success = mean(winner == "model_a" | winner == "tie"*0.5))

ggplot(model_ts, aes(date, success)) +
  geom_line() +
  geom_smooth(se = FALSE) +
  pub_theme() +
  labs(title = "Time Series Win Rate for Claude 3.5 Sonnet",
       x = "Date", y = "Success Rate")
```

\[insert discussion\]

Short list for your write-up (explain each briefly): Selection bias of prompts: The set of prompts users choose to run is not random. Users may cherry-pick prompts that favor particular models (e.g., models known to be strong at coding get coding prompts). Non-random assignment of pairings: Some models may be pitted against weaker opponents more often. Non-independent judges / vote stuffing: A small set of users could generate many matches and bias results (or coordinated voting). Presentation / ordering effects: The interface might show Model A on the left or highlight one model first; humans have position/anchoring biases. Anonymity revealed post-vote: If users learn model identity after voting, future votes may be biased by familiarity or brand. Ties/Both Bad handling: How ties are scored affects rankings (half-win vs. remove). Model versions change over time: The dataset aggregates across time; newer model versions likely better — mixing versions mixes abilities. Different prompt difficulty across categories: A model may appear strong simply because it faced easier prompts in a category. Judge expertise and calibration: Users differ in domain expertise; uncalibrated judgments (one user prefers different style) add measurement noise. How to game it: Create many accounts / sessions and push desired model to win. Submit prompts that expose known weaknesses in a competitor. Coordinate to flood-match a model against a weak competitor to inflate win-rate. Exploit UI order bias (always vote for left-side model).
