---
title: "Chatbot Ratings"
output: html_document
runtime: shiny
editor: visual
---

## Introduction

Chatbot Arena (lmarena.ai) is "an open-source platform for evaluating AI through human preference, developed by researchers at UC Berkeley [SkyLab](https://sky.cs.berkeley.edu/) and [LMSYS](https://lmsys.org/)." On the website <https://lmarena.ai/>, users can converse with two LLMs on any topic they choose. Users can then vote on which response they prefer (Model A or Model B) or can choose "Tie" or "Both bad" if they don't have a preference for either model. The LLMs are initially anonymous, but are revealed to the user after they vote.

## Data

The developers of this site provide data [here](https://huggingface.co/datasets/lmarena-ai/arena-human-preference-140k) on 140K conversations with pairwise human preferences. You can find this data in [this shared folder](https://yaleedu-my.sharepoint.com/:f:/g/personal/brian_macdonald_yale_edu/Eqo7cptmHAZAjNMv2F1dCj4BhZzs17e4hkfxbH8tMxLnwQ?e=Oo1DNN). There are 7 `.parquet` files. There is a README file in the GitHub repo that gives details about the data.

Each row is a matchup between two chat bots (e.g. Gemini 2.5 Pro vs Claude Sonnet 4), and the columns indicate a conversation ID, the names of the two models, who the user selected as the winner of the matchup, a common query that the two models were given, the category of the prompt (creative writing, math, code, etc.) and other information. See the README file for more details.

## Question

Which chatbot(s) appears to perform best in the eyes of the judges/users? In a matchup between any two models Model A and Model B, what are the outcome probabilities? How do these change if you focus on specific categories (e.g. writing, math, code, etc.) of prompts? Can you think of any reason the data may not be reliable, or any way this rating system could be "gamed"?

## Submission

Submit a PDF report with your code and analysis to Gradescope.

## Setup

```{r, eval=FALSE}
# read in data and bind rows
datalist <- vector(mode = "list", length = 7)

for(i in 0:6){
  data <- read_parquet(paste0("data/train-0000", i, "-of-00007.parquet")) %>% 
    as.data.frame() %>% 
    select(c(1:6, 11:14))
  datalist[[i+1]] <- data
  print(i)
}

df <- bind_rows(datalist)

flattened_data <- df %>% 
  unpack(cols = category_tag) %>%
  unpack(cols = c(creative_writing_v0.1, criteria_v0.1, if_v0.1, math_v0.1), 
         names_sep = "_")

saveRDS(flattened_data, file = "data.rds")
```

```{r}
# load packages
library(arrow)
library(dplyr)
library(tidyr)
library(stringr)
library(purrr)
library(ggplot2)
library(pubtheme)
library(tidyverse)
library(igraph)
library(ggraph)
library(tidygraph)
library(visNetwork)
library(scales)
library(viridisLite)
library(echarts4r)
library(BradleyTerry2)
library(shiny)
library(bslib)
library(tibble)
library(htmlwidgets)
library(DT)
library(rlang)

# load data
data <- readRDS("data.rds")
head(data)
colnames(data)
```

## Categories

```{r}
data <- data %>%
  mutate(
    # Convert weird values to logicals / numeric flags
    cw_flag   = creative_writing_v0.1_creative_writing %in% c(TRUE, "TRUE", 1),
    cw_score  = suppressWarnings(as.numeric(creative_writing_v0.1_score)),
    
    math_flag = math_v0.1_math %in% c(TRUE, "TRUE", 1),
    
    code_flag = is_code %in% c(TRUE, "TRUE", 1),
    
    if_flag   = if_v0.1_if %in% c(TRUE, "TRUE", 1),
    if_score  = suppressWarnings(as.numeric(if_v0.1_score)),
    
    # Build category
    category = case_when(
      code_flag ~ "code",
      math_flag ~ "math",
      if_flag | if_score > 0 ~ "reasoning",
      cw_flag | cw_score > 0 ~ "creative_writing",
      TRUE ~ "other"
    )
  )
```

## Best Chatbot (Shiny App)

Includes overall leaderboard and option to filter by category and match-up model.

```{r}
stopifnot(all(c("model_a","model_b","winner","category") %in% names(data)))

# ---------------- Helpers ----------------

# Expand each match into (model, category, win/tie) for both sides
# NOTE: For leaderboard & category profile we DO NOT count 'both_bad' as tie.
make_df2_leader <- function(df) {
  df %>%
    mutate(
      a_win = as.integer(winner == "model_a"),
      b_win = as.integer(winner == "model_b"),
      a_tie = as.integer(winner == "tie"),
      b_tie = a_tie
    ) %>%
    pivot_longer(c(model_a, model_b), names_to = "side", values_to = "model") %>%
    mutate(
      win = if_else(side == "model_a", a_win, b_win),
      tie = if_else(side == "model_a", a_tie, b_tie)
    )
}

# Wilson CI for proportion
wilson_ci <- function(successes, n, z = 1.96) {
  ifelse(n > 0, {
    phat <- successes / n
    denom <- 1 + (z^2)/n
    center <- (phat + (z^2)/(2*n)) / denom
    half   <- (z * sqrt((phat*(1-phat) + (z^2)/(4*n))/n)) / denom
    tibble(lower = pmax(0, center - half), upper = pmin(1, center + half))
  }, tibble(lower = NA_real_, upper = NA_real_))
}

all_categories <- sort(unique(data$category))

# ---------------- UI ----------------

ui <- fluidPage(
  theme = bs_theme(bootswatch = "flatly"),
  br(),
  titlePanel(div("Chatbot Performance and Ratings",
                 style = "padding-bottom: 20px;"
            )
  ),
  br(),
  sidebarLayout(
    sidebarPanel(
      class = "sticky-top",
      h4("Global Filters"),
      # View scope (Overall or one category)
      selectInput(
        "view_mode", "Category:",
        choices = c("Overall", all_categories),
        selected = "Overall"
      ),
      conditionalPanel(
        condition = "input.view_mode == 'Overall'",
        selectizeInput(
          "cat", "Filter Categories:",
          choices = all_categories, selected = all_categories,
          multiple = TRUE, options = list(plugins = list("remove_button"))
        )
      ),
      sliderInput("min_matches", "Minimum Matches", min = 1, max = 500, value = 30, step = 1)
    ),
    mainPanel(

      # ---- Leaderboard Card ----
      h3("Best Chatbots by Prompt Category"),
      div(
        style = "padding:20px; font-size:15px; line-height:1.5;",
        p("The leaderboard summarizes model performance across all categories by combining wins, ties, and losses (including 'both bad') into a composite success metric (Win + 0.5×Tie). Each model’s win rate and success rate are calculated from the full dataset of pairwise comparisons, filtering out models that participated in fewer than the minimum number of matches. Models are then ranked according to either win or success rate. Based on current data, the top five chatbots overall (excluding previews) are Gemini 2.5 Pro, Grok 4 0709, o3 2025-04-16, Chat-GPT-4o latest 20250326, and Deep Seek r1 0528. Across specific prompt categories, Gemini 2.5 Pro leads in coding tasks, Grok 4 0709 in math, Gemini 2.5 Pro preview 05-06 in reasoning, Gemini 2.5 Pro preview 05-06 in creative writing, and Gemini 2.5 Pro in 'other' tasks.")
      ),
      bslib::card(
        full_screen = TRUE, collapsible = TRUE, collapsed = FALSE,
        bslib::card_header(
          div(style="display:flex; align-items:center; justify-content:space-between; gap:16px; flex-wrap:wrap;",
              div(style="font-weight:600;", textOutput("main_title")),
              div(
                selectInput(
                  "sort_by", "Sort leaderboard by:",
                  choices = c("Success rate", "Win rate"),
                  selected = "Success rate", width = "220px"
                )
              )
          )
        ),
        bslib::card_body(
          div(style = "font-weight:600; font-size:16px; margin-bottom:8px;", 
              "Sorted Bar Chart"),
          echarts4rOutput("main_chart", height = "600px"),
          tags$hr(),
          div(style = "font-weight:600; margin:8px 0;", "Table of Rates"),
          DTOutput("main_table")
        )
      ),

      # ---- Category Profile Card ----
      br(),
      h3("Chatbot Strength & Weaknesses by Category"),
      div(
        style = "padding:20px; font-size:15px; line-height:1.5;",
        p("The category profile section measures how each chatbot performs within individual categories relative to the weighted category average. For the selected model, the bar chart shows per-category success rates against the mean of all models, while the radar plot normalizes these values to highlight relative strengths and weaknesses across categories This provides an in-depth view of how a single model’s capabilities vary across task types compared to the average.")
      ),
      bslib::card(
        full_screen = TRUE, collapsible = TRUE, collapsed = FALSE,
        bslib::card_header(
          div(style="display:flex; align-items:center; justify-content:space-between; gap:16px; flex-wrap:wrap;",
              div(style="font-weight:600;", textOutput("prof_title_bar")),
              div(
                selectInput("prof_model", "Selected model:", choices = NULL, width = "280px")
              )
          )
        ),
        bslib::card_body(
          # Card-local toggles
          div(style="display:flex; gap:20px; flex-wrap:wrap; align-items:center; margin-bottom:10px;",
              checkboxInput("prof_use_success", "Use Success Rate (Win + 0.5×Tie)", value = TRUE)
          ),
          tags$div(style = "font-weight:600; margin:8px 0;", "Sorted Bar Chart"),
          echarts4rOutput("prof_bar", height = "520px"),
          tags$hr(),
          tags$div(style = "font-weight:600; margin-bottom:6px;",
                   "Strengths & Weaknesses (Radar; normalized within selected categories)"),
          echarts4rOutput("prof_radar", height = "420px"),
          tags$hr(),
          tags$div(style = "font-weight:600; margin:8px 0;", "Table of Rates by Category"),
          DTOutput("prof_detail_table")
        )
      ),

      # ---- Bradley–Terry Card ----
      br(),
      h3("Estimated Match-Up Results by Category"),
      div(
        style = "padding:20px; font-size:15px; line-height:1.5;",
        p("In this section we use the Bradley–Terry model to estimate the latent ability of each chatbot based on head-to-head outcomes.The table and stacked bar chart show the probability that one model defeats another, including the empirical rate of ties or 'both-bad' outcomes. The heatmap summarizes these pairwise win probabilities across all models.")
      ),
      bslib::card(
        full_screen = TRUE, collapsible = TRUE, collapsed = FALSE,
        bslib::card_header(
          div(style="display:flex; align-items:center; justify-content:space-between; gap:16px; flex-wrap:wrap;",
              div(style="font-weight:600;", textOutput("bt_title")),
              div(
                selectInput("bt_model", "Selected model:",
                            choices = NULL, width = "320px")
              )
          )
        ),
        bslib::card_body(
          div(style = "font-weight:600; margin:8px 0;", "Head-to-Head Win Probabilities (Bradley–Terry)"),
          plotlyOutput("bt_heatmap", height = "680px"),
          tags$hr(),
          div(style = "font-weight:600; margin:8px 0;", "Sorted Bar Chart"),
          echarts4rOutput("bt_chart", height = "720px"),
          tags$hr(),
          div(style = "font-weight:600; margin:8px 0;", "Table of Predicted Rates"),
          DTOutput("bt_table")
        )
      )
    )
  )
)

# ---------------- SERVER ----------------

server <- function(input, output, session){

  # ----- Scoped data (shared by all cards) -----
  selected_df <- reactive({
    if (identical(input$view_mode, "Overall")) {
      df <- data
      if (!is.null(input$cat) && length(input$cat) > 0) {
        df <- df %>% filter(category %in% input$cat)
      }
      df
    } else {
      data %>% filter(category == input$view_mode)
    }
  })

  # All models available under current scope
  models_available <- reactive({
    df <- selected_df()
    df %>%
      select(model_a, model_b) %>%
      pivot_longer(everything(), values_to = "model") %>%
      distinct(model) %>%
      arrange(model) %>%
      pull(model)
  })

  # Keep profile & BT model dropdowns in sync with available models
  observeEvent(models_available(), {
    ms <- models_available()
    sel_prof <- if (length(ms) && input$prof_model %in% ms) input$prof_model else dplyr::first(ms)
    sel_bt   <- if (length(ms) && input$bt_model   %in% ms) input$bt_model   else dplyr::first(ms)
    updateSelectInput(session, "prof_model", choices = ms, selected = sel_prof)
    updateSelectInput(session, "bt_model",   choices = ms, selected = sel_bt)
  }, ignoreInit = FALSE)

  # ============ LEADERBOARD ============

  leaderboard <- reactive({
    df2 <- make_df2_leader(selected_df())
    sort_col <- if (identical(input$sort_by, "Win rate")) "win_rate" else "success_rate"

    df2 %>%
      group_by(model) %>%
      summarise(
        wins = sum(win),
        ties = sum(tie),
        matches = n(),
        win_rate = wins / matches,
        tie_rate = ties / matches,
        success_rate = (wins + 0.5 * ties) / matches,
        .groups = "drop"
      ) %>%
      filter(matches >= input$min_matches) %>%
      arrange(desc(.data[[sort_col]])) %>%
      mutate(
        rank     = row_number(),
        label    = sprintf("#%d  %s", rank, model),
        win_comp = round(win_rate, 3),
        tie_half = round(0.5 * tie_rate, 3)
      )
  })

  output$main_title <- renderText({
    scope <- if (identical(input$view_mode, "Overall")) {
      cats <- if (is.null(input$cat) || !length(input$cat)) "All categories"
      else paste(input$cat, collapse = ", ")
      paste0("Chatbots Leaderboard (Success Rate = Win Rate + 0.5×Tie Rate)")
    }
    scope
  })

  output$main_chart <- renderEcharts4r({
    lb <- leaderboard()
    validate(need(nrow(lb) > 0, "No models meet the minimum matches threshold."))

    metric_title <- if (identical(input$sort_by, "Win rate")) "Win Rate" else "Success Rate (Win + 0.5×Tie)"
    x_label      <- if (identical(input$sort_by, "Win rate")) "Win Rate" else "Success Rate"

    start_lab <- lb$label[1]
    end_lab   <- lb$label[min(10, nrow(lb))]

    lb |>
      e_charts(label) |>
      e_bar(win_comp, name = "Win Rate", stack = "rate",
            label = list(show = TRUE, position = "insideRight", color = "#fff", fontSize = 10)) |>
      e_bar(tie_half, name = "0.5*Tie Rate", stack = "rate",
            label = list(show = TRUE, position = "right", color = "#000", fontSize = 10)) |>
      e_flip_coords() |>
      e_y_axis(inverse = TRUE) |>
      e_legend(right = 10, top = 30) |>
      e_title(paste0("Chatbots Ranked by ", metric_title)) |>
      e_datazoom(y_index = 0, type = "slider", startValue = start_lab, endValue = end_lab) |>
      e_datazoom(y_index = 0, type = "inside", startValue = start_lab, endValue = end_lab) |>
      e_x_axis(max = 1,
               axisLabel = list(formatter = htmlwidgets::JS("function(x){return (x*100).toFixed(0)+'%';}"))) |>
      e_grid(left = 170) |>
      e_tooltip(show = FALSE) |>
      e_theme("infographic") |>
      e_toolbox_feature("saveAsImage")
  })

  output$main_table <- renderDT({
    lb <- leaderboard()
    validate(need(nrow(lb) > 0, "No models meet the minimum matches threshold."))
    tbl <- lb %>% select(model, wins, ties, matches, win_rate, tie_rate, success_rate)

    order_col <- if (identical(input$sort_by, "Win rate")) 4L else 6L
    
    tbl <- tibble(
      Model = tbl$model,
      Wins = tbl$wins,
      Ties = tbl$ties,
      Matches = tbl$matches,
      `Win Rate` = tbl$win_rate,
      `Tie Rate` = tbl$tie_rate,
      `Success Rate` = tbl$success_rate
    ) 

    datatable(
      tbl,
      rownames = FALSE,
      extensions = "Buttons",
      options = list(
        pageLength = 10,
        lengthMenu = c(5, 10, 25, 50, 100),
        order = list(list(order_col, "desc")),
        dom = "Bfrtip",
        buttons = c("copy", "csv", "excel")
      )
    ) %>%
      formatPercentage(columns = c("Win Rate", "Tie Rate", "Success Rate"), digits = 1)
  })

  # ============ CATEGORY PROFILE ============

  df2_profile <- reactive({
    make_df2_leader(selected_df())
  })

  model_cat_summary <- reactive({
    df2_profile() %>%
      group_by(model, category) %>%
      summarise(
        wins = sum(win),
        ties = sum(tie),
        matches = n(),
        win_rate = wins / matches,
        tie_rate = ties / matches,
        success_rate = (wins + 0.5 * ties) / matches,
        .groups = "drop"
      )
  })

  sel_profile <- reactive({
    metric <- if (isTRUE(input$prof_use_success)) "success_rate" else "win_rate"
    mcs <- model_cat_summary()

    me <- mcs %>%
      filter(model == input$prof_model) %>%
      rename(model_matches = matches)

    bench <- mcs %>%
      group_by(category) %>%
      summarise(
        bench = weighted.mean(.data[[metric]], w = .data$matches),
        total_matches = sum(.data$matches),
        .groups = "drop"
      )

    me %>%
      inner_join(bench, by = "category") %>%
      mutate(
        metric = .data[[metric]],
        lift = metric - bench
      ) %>%
      filter(model_matches >= input$min_matches) %>%
      arrange(desc(metric))
  })

  output$prof_title_bar <- renderText({
    scope <- if (identical(input$view_mode, "Overall")) {
      cats <- if (is.null(input$cat) || !length(input$cat)) "All categories"
      else paste(input$cat, collapse = ", ")
      paste0("Category Profile for ", input$prof_model)
    }
    scope
  })

  output$prof_bar <- renderEcharts4r({
    validate(
      need(length(models_available()) > 0, "No models in the current view/filter."),
      need(!is.null(input$prof_model) && nzchar(input$prof_model), "Select a model to view the profile.")
    )

    prof <- sel_profile() %>%
      filter(is.finite(metric), is.finite(bench)) %>%
      mutate(
        metric = ifelse(is.na(metric), 0, metric),
        bench  = ifelse(is.na(bench),  0, bench)
      )

    validate(need(nrow(prof) > 0, "No categories meet the minimum matches threshold."))

    chart_df <- prof %>%
      arrange(desc(metric)) %>%
      mutate(
        cat_lab = sprintf("%s  (n=%d)", category, model_matches),
        my_val  = round(metric, 6),
        bench_r = round(bench,  6)
      )

    chart_df |>
      e_charts(cat_lab) |>
      e_bar(
        my_val, name = input$prof_model,
        label = list(
          show = TRUE, position = "insideRight",
          color = "#fff", fontSize = 10
        )
      ) |>
      e_line(
        bench_r, 
        name = "Category Avg", 
        y_index = 0, 
        symbol = "circle", 
        symbolSize = 6,
        label = list(
               show = TRUE, position = 'right',
               color = '#000',fontSize = 10
             )) |>
      e_flip_coords() |>
      e_y_axis(inverse = TRUE) |>
      e_title("Model vs Category Average") |>
      e_x_axis(
        max = 1,
        axisLabel = list(formatter = htmlwidgets::JS("function(x){ return (x*100).toFixed(0)+'%'; }"))
      ) |>
      e_grid(left = 170) |>
      e_legend(right = 10, top = 30) |>
      e_datazoom(y_index = 0, type = "slider") |>
      e_theme("infographic") |>
      e_toolbox_feature("saveAsImage")
  })

  output$prof_radar <- renderEcharts4r({
    mcs <- model_cat_summary() %>% filter(matches >= input$min_matches)
    metric <- if (isTRUE(input$prof_use_success)) "success_rate" else "win_rate"

    cat_minmax <- mcs %>%
      group_by(category) %>%
      summarise(
        mn = suppressWarnings(min(.data[[metric]], na.rm = TRUE)),
        mx = suppressWarnings(max(.data[[metric]], na.rm = TRUE)),
        .groups = "drop"
      )

    me <- mcs %>%
      filter(model == input$prof_model) %>%
      inner_join(cat_minmax, by = "category") %>%
      mutate(
        val = case_when(
          is.finite(mx) & is.finite(mn) & mx > mn ~ (.data[[metric]] - mn) / (mx - mn),
          TRUE ~ 0.5
        )
      ) %>%
      arrange(category) %>%
      filter(is.finite(val))

    validate(need(nrow(me) >= 3, "Not enough categories with sufficient matches for radar."))

    df_wide <- me %>%
      select(category, val) %>%
      tidyr::pivot_wider(names_from = category, values_from = val) %>%
      mutate(across(everything(), ~ ifelse(is.na(.x), 0.5, .x))) %>%
      mutate(model = input$prof_model)

    cat_cols <- setdiff(names(df_wide), "model")
    validate(need(length(cat_cols) >= 3, "Radar needs ≥ 3 categories."))

    indicator_list <- lapply(cat_cols, function(nm) list(name = nm, max = 1))

    df_wide |>
      e_charts(model) |>
      e_radar(!!!rlang::syms(cat_cols), name = input$prof_model) |>
      e_radar_opts(indicator = indicator_list) |>
      e_title("Normalized Strengths by Category (0–1 within chosen categories)") |>
      e_tooltip(trigger = "item") |>
      e_theme("infographic")
  })

  output$prof_detail_table <- DT::renderDT({
    validate(
      need(length(models_available()) > 0, "No models in the current view/filter."),
      need(!is.null(input$prof_model) && nzchar(input$prof_model), "Select a model to view the profile.")
    )

    mcs <- model_cat_summary()
    metric_name <- if (isTRUE(input$prof_use_success)) "success_rate" else "win_rate"

    prof <- mcs %>%
      filter(model == input$prof_model) %>%
      rename(model_matches = matches)

    bench <- mcs %>%
      group_by(category) %>%
      summarise(
        bench = weighted.mean(.data[[metric_name]], w = .data$matches),
        total_matches = sum(.data$matches),
        .groups = "drop"
      )

    prof <- prof %>%
      inner_join(bench, by = "category") %>%
      mutate(
        metric = .data[[metric_name]],
        lift = metric - bench
      ) %>%
      filter(model_matches >= input$min_matches) %>%
      arrange(desc(metric))

    validate(need(nrow(prof) > 0, "No categories meet the minimum matches threshold."))

    if (isTRUE(input$prof_use_success)) {
      ci <- wilson_ci(successes = prof$wins + 0.5 * prof$ties, n = prof$model_matches)
    } else {
      ci <- wilson_ci(successes = prof$wins, n = prof$model_matches)
    }

    tbl <- tibble(
      Category = prof$category,
      `Matches (model)` = prof$model_matches,
      Wins = prof$wins,
      Ties = prof$ties,
      Rate = prof$metric,
      `Category Avg` = prof$bench,
      `Lift (pts)` = round(100 * prof$lift, 1),
      `Matches (category total)` = prof$total_matches
    )

    DT::datatable(
      tbl,
      rownames = FALSE,
      extensions = "Buttons",
      options = list(
        pageLength = 10,
        lengthMenu = c(5, 10, 25, 50, 100),
        order = list(list(4, "desc")), # Rate
        dom = "Bfrtip",
        buttons = c("copy", "csv", "excel")
      )
    ) %>%
      DT::formatPercentage(columns = c("Rate", "Category Avg"), digits = 1)
  })

  # ============ BRADLEY–TERRY (decisive-only fit) ============

  # Pair counts including ties & both_bad for per-pair filtering and for p_tie_or_both_bad
  bt_pair_counts <- reactive({
    selected_df() %>%
      transmute(
        A = pmin(model_a, model_b),
        B = pmax(model_a, model_b),
        is_tie_or_bb = as.integer(winner %in% c("tie","both_bad"))
      ) %>%
      filter(A != B) %>%
      summarise(
        matches = dplyr::n(),
        ties_or_bb = sum(is_tie_or_bb),
        .by = c(A, B)
      )
  })

  # Safe BT fit on decisive matches (no ties/both_bad in the likelihood)
  bt_fit <- reactive({
    df <- selected_df() %>%
      filter(winner %in% c("model_a", "model_b")) %>%
      transmute(
        player1 = model_a,
        player2 = model_b,
        r1 = as.numeric(winner == "model_a"),
        r2 = as.numeric(winner == "model_b")
      )

    validate(need(nrow(df) > 0, "No decisive matches in the selected filter."))

    players <- sort(unique(c(df$player1, df$player2)))
    df <- df %>% mutate(
      player1 = factor(player1, levels = players),
      player2 = factor(player2, levels = players)
    )

    fit <- tryCatch(
      BTm(cbind(r1, r2), player1, player2, data = df, id = "player"),
      error = function(e) NULL
    )
    validate(need(!is.null(fit), "BT model failed to fit for the selected filter (disconnected or separable)."))
    fit
  })

  bt_abilities <- reactive({
      fit <- bt_fit()
      ab <- BTabilities(fit) %>% as.data.frame() %>% tibble::rownames_to_column("model") %>% select(model, ability)
      setNames(ab$ability, ab$model)
    })
    
  output$bt_heatmap <- renderPlotly({
    # Abilities (named numeric vector) and available models in scope
    ab_map   <- bt_abilities()                        # named numeric vector: ability by model
    choices  <- models_available()                    # models visible in current view/filter
  
    validate(need(length(choices) > 1, "Need at least two models to draw the heatmap."))
  
    # Put abilities into a data frame
    abilities <- tibble(model = names(ab_map), ability = as.numeric(ab_map))
  
    # Order by leaderboard success_rate (already sorted there)
    lb <- leaderboard()
    validate(need(nrow(lb) > 0, "No leaderboard rows to drive model ordering."))
  
    model_order <- intersect(lb$model, abilities$model)
    validate(need(length(model_order) > 1, "Not enough overlap between BT models and leaderboard models."))
  
    # Full grid of pairs with P(row beats col) from BT
    grid <- expand.grid(
      row_model = model_order,
      col_model = model_order,
      stringsAsFactors = FALSE
    ) %>%
      mutate(
        prob = ifelse(
          row_model == col_model, NA_real_,
          plogis(ab_map[row_model] - ab_map[col_model])
        )
      )
  
    # Factors to enforce ordering (rows: top=best; cols: left=best)
    grid$row_model <- factor(grid$row_model, levels = rev(model_order))
    grid$col_model <- factor(grid$col_model, levels = model_order)
  
    # Hover text
    grid <- grid %>% mutate(
      hover = sprintf("Row: %s<br>Col: %s<br>P(Row beats Col): %s",
                      row_model, col_model,
                      ifelse(is.na(prob), "—", percent(prob, 0.1)))
    )
  
    p <- ggplot(grid, aes(x = col_model, y = row_model, fill = prob, text = hover)) +
      geom_tile(na.rm = FALSE) +
      scale_fill_viridis_c(
        option = "D", limits = c(0, 1),
        labels = percent_format(accuracy = 1),
        name = "P(Row beats Col)"
      ) +
      scale_color_identity() +
      scale_linewidth_identity() +
      coord_equal() +
      labs(
        title = "Head-to-Head Win Probabilities (Bradley–Terry)",
        x = "Column model",
        y = "Row model"
      ) +
      theme_minimal(base_size = 13) +
      theme(
        plot.title.position = "plot",
        axis.title.x = element_text(margin = margin(t = 0)),  # reduce gap below plot
        axis.title.y = element_text(margin = margin(r = 0)),
        axis.text.x = element_text(size = 3, angle = 45, hjust = 1, vjust = 1),
        axis.text.y = element_text(size = 4),
        panel.grid = element_blank()
      )
    
    ggplotly(p, tooltip = "text")
  })

  # Build predicted rates table for the selected model
  bt_make_vs <- function(sel, model_choices, ab_map, pair_counts) {
    opps <- setdiff(model_choices, sel)
    if (!(sel %in% names(ab_map)) || !length(opps)) {
      return(tibble(opponent=character(), matches_vs_selected=integer(),
                    p_sel_vs_row=numeric(), p_row_vs_sel=numeric(), p_tie_or_both_bad=numeric()))
    }

    base <- tibble(opponent = opps) %>%
      mutate(
        p_sel_vs_row = ifelse(opponent %in% names(ab_map),
                              plogis(ab_map[sel] - ab_map[opponent]), NA_real_),
        p_row_vs_sel = ifelse(opponent %in% names(ab_map),
                              plogis(ab_map[opponent] - ab_map[sel]), NA_real_),
        A = pmin(sel, opponent), B = pmax(sel, opponent)
      ) %>%
      left_join(pair_counts, by = c("A","B")) %>%
      mutate(
        matches_vs_selected = round(replace_na(matches, 0L),3),
        # empirical tie/both_bad rate for this unordered pair
        p_tie_or_both_bad = round(replace_na(ties_or_bb / pmax(matches, 1), 0),3),
        p_sel_vs_row = round(replace_na(p_sel_vs_row, 0),3),
        p_row_vs_sel = round(replace_na(p_row_vs_sel, 0),3)
      ) %>%
      select(opponent, matches_vs_selected, p_sel_vs_row, p_row_vs_sel, p_tie_or_both_bad)

    # Keep rows with enough direct matches vs the selected model
    base %>%
      filter(matches_vs_selected >= input$min_matches) %>%
      arrange(desc(p_row_vs_sel))
  }

  output$bt_title <- renderText({
    cats_txt <- if (identical(input$view_mode, "Overall")) {
      cats <- if (is.null(input$cat) || !length(input$cat)) "All categories"
      else paste(input$cat, collapse = ", ")
      paste0("One-on-One Match-Up Predicted Win Probabilities Against ", input$bt_model)
    }
    cats_txt
  })

  output$bt_chart <- renderEcharts4r({
    ab_map   <- bt_abilities()
    pair_cnt <- bt_pair_counts()
    choices  <- models_available()
    validate(need(length(choices) > 0, "No models available for the selected filter."))

    df_all <- bt_make_vs(input$bt_model, choices, ab_map, pair_cnt)
    validate(need(nrow(df_all) > 0, "No head-to-head predictions meet the min-matches threshold."))

    # Rank labels and default zoom to top-10
    df <- df_all %>%
      mutate(rank = row_number(), label = sprintf("#%d  %s", rank, opponent))
    start_lab <- df$label[1]
    end_lab   <- df$label[min(10, nrow(df))]

    # Stacked bars: p(row beats selected) + p(tie_or_both_bad)
    df |>
      e_charts(label) |>
      e_bar(p_row_vs_sel, name = "P(Row beats Selected)", stack = "prob",
            label = list(show = TRUE, position = "insideRight", color = "#fff", fontSize = 10
                         )
            ) |>
      e_bar(p_tie_or_both_bad, name = "P(Tie or Both-bad)", stack = "prob",
            label = list(show = TRUE, position = "right", color = "#000", fontSize = 10
                         )
            ) |>
      e_flip_coords() |>
      e_y_axis(inverse = TRUE) |>
      e_legend(right = 10, top = 30) |>
      e_title(paste0("Predicted Probability of Winning Against ", input$bt_model)) |>
      e_datazoom(y_index = 0, type = "slider", startValue = start_lab, endValue = end_lab) |>
      e_datazoom(y_index = 0, type = "inside", startValue = start_lab, endValue = end_lab) |>
      e_x_axis(max = 1,
               axisLabel = list(formatter = htmlwidgets::JS("function(x){return (x*100).toFixed(0)+'%';}"))) |>
      e_grid(left = 240) |>
      e_tooltip(show = FALSE) |>
      e_theme("infographic") |>
      e_toolbox_feature("saveAsImage")
  })

  output$bt_table <- DT::renderDT({
    ab_map   <- bt_abilities()
    pair_cnt <- bt_pair_counts()
    choices  <- models_available()
    validate(need(length(choices) > 0, "No models available for the selected filter."))

    tbl <- bt_make_vs(input$bt_model, choices, ab_map, pair_cnt)
    validate(need(nrow(tbl) > 0, "No head-to-head rows meet the min-matches threshold."))

    tbl <- tibble(
      Opponent = tbl$opponent,
      `# Matches` = tbl$matches_vs_selected,
      `P(Selected Model Wins)` = tbl$p_sel_vs_row,
      `P(Opponent Wins)` = tbl$p_row_vs_sel,
      `P(Tie or Both Bad)` = tbl$p_tie_or_both_bad
    ) 
    
    DT::datatable(
      tbl,
      rownames = FALSE,
      extensions = "Buttons",
      options = list(
        pageLength = 10,
        lengthMenu = c(5, 10, 25, 50, 100),
        order = list(list(3, "desc")),  # p_row_vs_sel
        dom = "Bfrtip",
        buttons = c("copy", "csv", "excel")
      )
    ) %>%
      DT::formatPercentage(columns = c("P(Selected Model Wins)", "P(Opponent Wins)", "P(Tie or Both Bad)"), digits = 1)
  })
}

shinyApp(ui, server)
```

## Limitations

```{r}
# Count matches per pair and per model — sparsity:
df2 %>% count(model_a, model_b) %>% arrange(n)  # look for tiny n
model_summary %>% arrange(matches)  # models with very few matches

# votes per session
session_votes <- flat_data %>%
  count(evaluation_session_id)

ggplot(session_votes, aes(n)) +
  geom_histogram(bins = 40) +
  pub_theme() +
  labs(title = "Votes per Session",
       x = "Votes in Session", y = "Count of Sessions")

# Distribution of judges / sessions: are a small set of users contributing many votes? If you have evaluation_session_id or timestamp, check duplicates:
df2 %>% count(evaluation_session_id) %>% top_n(20, n)

# Order effect (evaluation_order): does the first-shown model get more votes?
df2 %>% group_by(evaluation_order) %>%
  summarise(win_rate_first = mean(if_else(evaluation_order == "A", a_win, b_win)))

# Time trends: models have versions; check timestamps — a model may improve over time.
library(lubridate)
df2 %>% mutate(ts = ymd_hms(timestamp)) %>% arrange(ts) %>% 
  NULL

model_ts <- flat_data %>%
  filter(model_a == "Claude 3.5 Sonnet") %>%   # example — you can parametrize this
  mutate(date = as.Date(timestamp)) %>%
  group_by(date) %>%
  summarise(success = mean(winner == "model_a" | winner == "tie"*0.5))

ggplot(model_ts, aes(date, success)) +
  geom_line() +
  geom_smooth(se = FALSE) +
  pub_theme() +
  labs(title = "Time Series Win Rate for Claude 3.5 Sonnet",
       x = "Date", y = "Success Rate")
```

\[insert discussion\]

Short list for your write-up (explain each briefly): Selection bias of prompts: The set of prompts users choose to run is not random. Users may cherry-pick prompts that favor particular models (e.g., models known to be strong at coding get coding prompts). Non-random assignment of pairings: Some models may be pitted against weaker opponents more often. Non-independent judges / vote stuffing: A small set of users could generate many matches and bias results (or coordinated voting). Presentation / ordering effects: The interface might show Model A on the left or highlight one model first; humans have position/anchoring biases. Anonymity revealed post-vote: If users learn model identity after voting, future votes may be biased by familiarity or brand. Ties/Both Bad handling: How ties are scored affects rankings (half-win vs. remove). Model versions change over time: The dataset aggregates across time; newer model versions likely better — mixing versions mixes abilities. Different prompt difficulty across categories: A model may appear strong simply because it faced easier prompts in a category. Judge expertise and calibration: Users differ in domain expertise; uncalibrated judgments (one user prefers different style) add measurement noise. How to game it: Create many accounts / sessions and push desired model to win. Submit prompts that expose known weaknesses in a competitor. Coordinate to flood-match a model against a weak competitor to inflate win-rate. Exploit UI order bias (always vote for left-side model).
