---
title: "Chatbot Ratings"
output: html_document
runtime: shiny
editor: visual
---

## Introduction

Chatbot Arena (lmarena.ai) is "an open-source platform for evaluating AI through human preference, developed by researchers at UC Berkeley [SkyLab](https://sky.cs.berkeley.edu/) and [LMSYS](https://lmsys.org/)." On the website <https://lmarena.ai/>, users can converse with two LLMs on any topic they choose. Users can then vote on which response they prefer (Model A or Model B) or can choose "Tie" or "Both bad" if they don't have a preference for either model. The LLMs are initially anonymous, but are revealed to the user after they vote.

## Data

The developers of this site provide data [here](https://huggingface.co/datasets/lmarena-ai/arena-human-preference-140k) on 140K conversations with pairwise human preferences. You can find this data in [this shared folder](https://yaleedu-my.sharepoint.com/:f:/g/personal/brian_macdonald_yale_edu/Eqo7cptmHAZAjNMv2F1dCj4BhZzs17e4hkfxbH8tMxLnwQ?e=Oo1DNN). There are 7 `.parquet` files. There is a README file in the GitHub repo that gives details about the data.

Each row is a matchup between two chat bots (e.g. Gemini 2.5 Pro vs Claude Sonnet 4), and the columns indicate a conversation ID, the names of the two models, who the user selected as the winner of the matchup, a common query that the two models were given, the category of the prompt (creative writing, math, code, etc.) and other information. See the README file for more details.

## Question

Which chatbot(s) appears to perform best in the eyes of the judges/users? In a matchup between any two models Model A and Model B, what are the outcome probabilities? How do these change if you focus on specific categories (e.g. writing, math, code, etc.) of prompts? Can you think of any reason the data may not be reliable, or any way this rating system could be "gamed"?

## Submission

Submit a PDF report with your code and analysis to Gradescope.

## Setup

```{r, eval=FALSE}
# read in data and bind rows
datalist <- vector(mode = "list", length = 7)

for(i in 0:6){
  data <- read_parquet(paste0("data/train-0000", i, "-of-00007.parquet")) %>% 
    as.data.frame() %>% 
    select(c(1:6, 11:14))
  datalist[[i+1]] <- data
  print(i)
}

df <- bind_rows(datalist)

flattened_data <- df %>% 
  unpack(cols = category_tag) %>%
  unpack(cols = c(creative_writing_v0.1, criteria_v0.1, if_v0.1, math_v0.1), 
         names_sep = "_")

saveRDS(flattened_data, file = "data.rds")
```

```{r}
# load packages
library(arrow)
library(dplyr)
library(tidyr)
library(stringr)
library(purrr)
library(ggplot2)
library(pubtheme)
library(tidyverse)
library(igraph)
library(ggraph)
library(tidygraph)
library(visNetwork)
library(scales)
library(viridisLite)
library(echarts4r)
library(BradleyTerry2)

# load data
data <- readRDS("data.rds")
head(data)
colnames(data)
```

## Categories

```{r}
data <- data %>%
  mutate(
    # Convert weird values to logicals / numeric flags
    cw_flag   = creative_writing_v0.1_creative_writing %in% c(TRUE, "TRUE", 1),
    cw_score  = suppressWarnings(as.numeric(creative_writing_v0.1_score)),
    
    math_flag = math_v0.1_math %in% c(TRUE, "TRUE", 1),
    
    code_flag = is_code %in% c(TRUE, "TRUE", 1),
    
    if_flag   = if_v0.1_if %in% c(TRUE, "TRUE", 1),
    if_score  = suppressWarnings(as.numeric(if_v0.1_score)),
    
    # Build category
    category = case_when(
      code_flag ~ "code",
      math_flag ~ "math",
      if_flag | if_score > 0 ~ "reasoning",
      cw_flag | cw_score > 0 ~ "creative_writing",
      TRUE ~ "other"
    )
  )
count(data, category, sort = TRUE)
```

## Best Chatbot

```{r}
df2 <- data %>%
  mutate(
    a_win = case_when(winner == "model_a" ~ 1, TRUE ~ 0),
    b_win = case_when(winner == "model_b" ~ 1, TRUE ~ 0),
    a_tie = case_when(winner == "tie" ~ 1, TRUE ~ 0),
    b_tie = a_tie
  ) %>%
  pivot_longer(
    cols = c(model_a, model_b),
    names_to = "side",
    values_to = "model"
  ) %>%
  mutate(
    win = if_else(side == "model_a", a_win, b_win),
    tie = if_else(side == "model_a", a_tie, b_tie)
  )

model_summary <- df2 %>%
  group_by(model) %>%
  summarise(
    wins = sum(win),
    ties = sum(tie),
    matches = n(), 
    win_rate = wins / matches,
    tie_rate = ties / matches,
    success_rate = (wins + 0.5 * ties) / matches,
    .groups = "drop"
  ) %>%
  arrange(desc(success_rate))

model_summary

# chart
leaderboard <- model_summary %>%
  arrange(desc(success_rate)) %>%
  mutate(
    rank     = row_number(),
    label    = sprintf("#%d  %s", rank, model),
    win_comp = round(win_rate,3),
    tie_half = round(0.5 * tie_rate,3)
  )

top10_start <- leaderboard$label[1]
top10_end   <- leaderboard$label[min(10, nrow(leaderboard))]

leaderboard |>
  e_charts(label) |>
  e_bar(
    win_comp, name = "Win Rate", stack = "rate",
    label = list(
      show = TRUE, position = "insideRight",
      color = "#fff", fontSize = 10)
  ) |>
  e_bar(
    tie_half, name = "0.5*Tie Rate", stack = "rate",
    label = list(
      show = TRUE, position = "insideRight",
      color = "#fff", fontSize = 10)
  ) |>
  e_flip_coords() |>
  e_y_axis(inverse = TRUE) |>
  e_legend(orient = "vertical", right = 10, top = 40) |>
  e_title("Chatbots Ranked by Success Rate (Win + 0.5*Tie)") |>
  e_datazoom(y_index = 0, type = "slider",
             startValue = top10_start, endValue = top10_end) |>
  e_datazoom(y_index = 0, type = "inside",
             startValue = top10_start, endValue = top10_end) |>
  e_x_axis(max = 1,
           axisLabel = list(formatter = htmlwidgets::JS("function(x){return (x*100).toFixed(0)+'%';}"))) |>
  e_axis_labels(y = "Model (ranked)", x = "Success Rate") |>
  e_grid(left = 170) |>
  e_tooltip(show = FALSE) |>
  e_theme("infographic") |>
  e_toolbox_feature("saveAsImage")
```

## Best Chatbot (Shiny App)

Includes overall leaderboard and option to filter by category.

```{r}
library(shiny)
library(bslib)
library(dplyr)
library(tidyr)
library(tibble)
library(echarts4r)
library(htmlwidgets)

all_categories <- sort(unique(data$category))

ui <- page_fillable(
  theme = bs_theme(bootswatch = "flatly"),
  layout_sidebar(
    sidebar = sidebar(
      selectizeInput("cat", "Filter categories (for Overall view):",
                     choices = all_categories, selected = all_categories,
                     multiple = TRUE, options = list(plugins = list("remove_button"))),
      sliderInput("min_matches", "Min matches", min = 1, max = 500, value = 30, step = 1),
      selectInput("leader_cat", "By-category leaderboard:", choices = c(all_categories), selected = all_categories[1])
    ),
    card(
      full_screen = TRUE, collapsible = TRUE,
      card_header("Overall Leaderboard (Win + 0.5×Tie)"),
      card_body(echarts4rOutput("leaders_all", height = "600px"))
    ),
    card(
      full_screen = TRUE, collapsible = TRUE,
      card_header(textOutput("leaders_cat_title")),
      card_body(echarts4rOutput("leaders_cat", height = "600px"))
    )
  )
)

server <- function(input, output, session){

  filtered_data_overall <- reactive({
    df <- data
    if (!is.null(input$cat) && length(input$cat) > 0) df <- df %>% filter(category %in% input$cat)
    df
  })

  make_df2 <- function(df) {
    df %>%
      mutate(
        a_win = as.integer(winner == "model_a"),
        b_win = as.integer(winner == "model_b"),
        a_tie = as.integer(winner == "tie"),
        b_tie = a_tie
      ) %>%
      pivot_longer(c(model_a, model_b), names_to = "side", values_to = "model") %>%
      mutate(
        win = if_else(side == "model_a", a_win, b_win),
        tie = if_else(side == "model_a", a_tie, b_tie)
      )
  }

  leaderboard_overall <- reactive({
    df2 <- make_df2(filtered_data_overall())
    df2 %>%
      group_by(model) %>%
      summarise(
        wins = sum(win),
        ties = sum(tie),
        matches = n(),
        win_rate = wins / matches,
        tie_rate = ties / matches,
        success_rate = (wins + 0.5 * ties) / matches,
        .groups = "drop"
      ) %>%
      filter(matches >= input$min_matches) %>%
      arrange(desc(success_rate)) %>%
      mutate(
        rank     = row_number(),
        label    = sprintf("#%d  %s", rank, model),
        win_comp = round(win_rate, 3),
        tie_half = round(0.5 * tie_rate, 3)
      )
  })

  output$leaders_all <- renderEcharts4r({
    lb <- leaderboard_overall()
    validate(need(nrow(lb) > 0, "No models meet the minimum matches threshold."))
    start_lab <- lb$label[1]
    end_lab   <- lb$label[min(10, nrow(lb))]

    lb |>
      e_charts(label) |>
      e_bar(win_comp, name = "Win Rate", stack = "rate",
            label = list(show = TRUE, position = "insideRight", color = "#fff", fontSize = 10)) |>
      e_bar(tie_half, name = "0.5*Tie Rate", stack = "rate",
            label = list(show = TRUE, position = "insideRight", color = "#fff", fontSize = 10)) |>
      e_flip_coords() |>
      e_y_axis(inverse = TRUE) |>
      e_legend(orient = "vertical", right = 10, top = 40) |>
      e_title("Chatbots Ranked by Success Rate (Win + 0.5×Tie)") |>
      e_datazoom(y_index = 0, type = "slider", startValue = start_lab, endValue = end_lab) |>
      e_datazoom(y_index = 0, type = "inside", startValue = start_lab, endValue = end_lab) |>
      e_x_axis(max = 1,
               axisLabel = list(formatter = htmlwidgets::JS("function(x){return (x*100).toFixed(0)+'%';}"))) |>
      e_axis_labels(y = "Model (ranked)", x = "Success Rate") |>
      e_grid(left = 170) |>
      e_tooltip(show = FALSE) |>
      e_theme("infographic") |>
      e_toolbox_feature("saveAsImage")
  })

  leaderboard_by_cat <- reactive({
    req(input$leader_cat)
    df_cat <- data %>% filter(category == input$leader_cat)
    df2 <- make_df2(df_cat)
    df2 %>%
      group_by(model) %>%
      summarise(
        wins = sum(win),
        ties = sum(tie),
        matches = n(),
        win_rate = wins / matches,
        tie_rate = ties / matches,
        success_rate = (wins + 0.5 * ties) / matches,
        .groups = "drop"
      ) %>%
      filter(matches >= input$min_matches) %>%
      arrange(desc(success_rate)) %>%
      mutate(
        rank     = row_number(),
        label    = sprintf("#%d  %s", rank, model),
        win_comp = round(win_rate, 3),
        tie_half = round(0.5 * tie_rate, 3)
      )
  })

  output$leaders_cat_title <- renderText({
    paste0("By-Category Leaderboard — ", input$leader_cat,
           " (Win + 0.5×Tie; min matches = ", input$min_matches, ")")
  })

  output$leaders_cat <- renderEcharts4r({
    lb <- leaderboard_by_cat()
    validate(need(nrow(lb) > 0, "No models meet the minimum matches threshold for this category."))
    start_lab <- lb$label[1]
    end_lab   <- lb$label[min(10, nrow(lb))]

    lb |>
      e_charts(label) |>
      e_bar(win_comp, name = "Win Rate", stack = "rate",
            label = list(show = TRUE, position = "insideRight", color = "#fff", fontSize = 10)) |>
      e_bar(tie_half, name = "0.5*Tie Rate", stack = "rate",
            label = list(show = TRUE, position = "insideRight", color = "#fff", fontSize = 10)) |>
      e_flip_coords() |>
      e_y_axis(inverse = TRUE) |>
      e_legend(orient = "vertical", right = 10, top = 40) |>
      e_title(paste0("Chatbots in ", input$leader_cat, " Ranked by Success Rate")) |>
      e_datazoom(y_index = 0, type = "slider", startValue = start_lab, endValue = end_lab) |>
      e_datazoom(y_index = 0, type = "inside", startValue = start_lab, endValue = end_lab) |>
      e_x_axis(max = 1,
               axisLabel = list(formatter = htmlwidgets::JS("function(x){return (x*100).toFixed(0)+'%';}"))) |>
      e_axis_labels(y = "Model (ranked)", x = "Success Rate") |>
      e_grid(left = 170) |>
      e_tooltip(show = FALSE) |>
      e_theme("infographic") |>
      e_toolbox_feature("saveAsImage")
  })
}

shinyApp(ui, server)

```

## Match-Up Outcome Probabilities

```{r}
# Filter out ties for standard BT model (ties need a separate extension)
bt_data <- data %>%
  filter(winner %in% c("model_a", "model_b")) %>%
  transmute(
    player1 = model_a,
    player2 = model_b,
    result1 = as.numeric(winner == "model_a"),
    result2 = as.numeric(winner == "model_b")
  )
# ensure same factor levels for both sides
players <- sort(unique(c(bt_data$player1, bt_data$player2)))
bt_data <- bt_data %>%
  mutate(
    player1 = factor(player1, levels = players),
    player2 = factor(player2, levels = players)
  )

bt_model <- BTm(
  cbind(result1, result2), player1, player2,
  data = bt_data, id = "player"
)

summary(bt_model)

abilities <- BTabilities(bt_model) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("model") %>%
  arrange(desc(ability)) %>%
  mutate(
    strength = exp(ability),        # convert to positive “skill” scale
    normalized = strength / sum(strength)
  )

head(abilities, 10)

predict_bt <- function(modelA, modelB, abilities_df = abilities) {
  a <- abilities_df$ability[abilities_df$model == modelA]
  b <- abilities_df$ability[abilities_df$model == modelB]
  pA <- exp(a) / (exp(a) + exp(b))
  tibble(modelA, modelB, P_A_wins = pA, P_B_wins = 1 - pA)
}
```

```{r}
# 1) order by success_rate (highest → lowest)
model_order <- model_summary %>% arrange(desc(success_rate)) %>% pull(model)

# keep only models present in the BT abilities table
model_order <- intersect(model_order, abilities$model)

# 2) ability lookup
ab_map <- setNames(abilities$ability, abilities$model)

# 3) full grid of pairs with BT prob P(row beats col)
grid <- expand.grid(
  row_model = model_order,
  col_model = model_order,
  stringsAsFactors = FALSE
) %>%
  mutate(
    prob = ifelse(
      row_model == col_model, NA_real_,
      plogis(ab_map[row_model] - ab_map[col_model])  # exp(a)/(exp(a)+exp(b))
    )
  )

# factors to get the desired ordering in the plot:
grid$row_model <- factor(grid$row_model, levels = rev(model_order))  # rows: top = best
grid$col_model <- factor(grid$col_model, levels = model_order)       # cols: left = best

library(plotly)
library(scales)

grid <- grid %>% mutate(
  hover = sprintf("Row: %s<br>Col: %s<br>P(Row beats Col): %s",
                  row_model, col_model,
                  ifelse(is.na(prob), "—", percent(prob, 0.1)))
)

p <- ggplot(grid, aes(x = col_model, y = row_model, fill = prob, text = hover)) +
  geom_tile(color = "white", linewidth = 0.2, na.rm = FALSE) +
  scale_fill_viridis_c(option = "D", limits = c(0,1),
                       labels = percent_format(accuracy = 1),
                       name = "P(Row beats Col)") +
  coord_equal() +
  labs(title = "Head-to-Head Win Probabilities (Bradley–Terry)",
       x = "Column model", y = "Row model") +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(size = 4, angle = 45, hjust = 1, vjust = 1),
        axis.text.y = element_text(size = 4))

ggplotly(p, tooltip = "text")

```

```{r}
# edges and nodes
pairs <- data %>% 
  count(model_a, model_b, name = "matches") %>%
  mutate(
    edge_width = log1p(matches),   # damp outliers for thickness
    match_bin = cut(                # edge COLOR uses raw-match bins (for legend)
      matches,
      breaks = c(1,5,10,25,50,100,Inf),
      labels = c("1–5","6–10","11–25","26–50","51–100","100+"),
      right = TRUE, include.lowest = TRUE
    )
  )

nodes <- model_summary %>% 
  select(model, success_rate)

g <- graph_from_data_frame(d = pairs, vertices = nodes, directed = FALSE)

# ---- ADDED: total matches per model (sum of incident edge matches) ----------
totals <- pairs %>%
  select(model_a, model_b, matches) %>%
  pivot_longer(c(model_a, model_b), names_to = "side", values_to = "model") %>%
  group_by(model) %>%
  summarise(total_matches = sum(matches, na.rm = TRUE), .groups = "drop")

# success rate bins for legend + node popups
nodes2 <- model_summary %>%
  select(model, success_rate) %>%
  rename(id = model) %>%
  left_join(totals, by = c("id" = "model")) %>%     # ADDED
  mutate(
    success_bin = cut(
      success_rate,
      breaks = seq(min(success_rate, na.rm = TRUE),  # ADDED: na.rm for safety
                   max(success_rate, na.rm = TRUE),
                   length.out = 7),
      include.lowest = TRUE
    ),
    color = viridisLite::viridis(6)[as.integer(success_bin)],
    label = id,
    font.color = color,
    # ADDED: tooltip content
    title = paste0(
      "<b>", id, "</b><br>",
      "Success rate: ", percent(success_rate, accuracy = 0.1), "<br>",
      "Total matches: ", ifelse(is.na(total_matches), 0, total_matches)
    )
  )

# --- NEW: empirical head-to-head outcome probabilities per unordered pair ----
edge_probs <- data %>%
  transmute(
    A = pmin(model_a, model_b),
    B = pmax(model_a, model_b),
    res = case_when(
      winner == "tie" ~ "tie",
      winner == "model_a" & model_a <= model_b ~ "A_win",
      winner == "model_b" & model_a  > model_b ~ "A_win",
      winner == "model_b" & model_a <= model_b ~ "B_win",
      winner == "model_a" & model_a  > model_b ~ "B_win",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(res), A != B) %>%
  count(A, B, res) %>%
  tidyr::pivot_wider(
    names_from = res, values_from = n, values_fill = 0
  ) %>%
  mutate(
    total = A_win + B_win + tie,
    pA = ifelse(total > 0, A_win / total, NA_real_),  # P(A beats B)
    pB = ifelse(total > 0, B_win / total, NA_real_),  # P(B beats A)
    pT = ifelse(total > 0, tie   / total, NA_real_)   # tie rate
  )

# --- NEW: join probabilities onto your existing pairs in a direction-aware way
pairs_with_probs <- pairs %>%
  mutate(A = pmin(model_a, model_b),
         B = pmax(model_a, model_b)) %>%
  left_join(edge_probs, by = c("A","B")) %>%
  mutate(
    from = model_a, to = model_b,
    p_from = if_else(model_a <= model_b, pA, pB),
    p_to   = if_else(model_a <= model_b, pB, pA),
    tie_rate = pT
  )

# darker greys, first one darker manually (keep your palette)
edge_palette <- RColorBrewer::brewer.pal(6, "Greys")
edge_palette[1] <- "#6f6f6f"

vn_edges2 <- pairs_with_probs %>%
  transmute(
    from, to,
    width = pmax(1, round(log1p(matches))),
    color = edge_palette[as.integer(match_bin)],
    # Edge label now includes matchups + probabilities (rounded to 2dp)
    title = sprintf(
      "<b>%s ↔ %s</b><br>Matches: %d<br>%s win prob: %.2f%%<br>%s win prob: %.2f%%%s",
      from, to, matches,
      from, 100 * p_from,
      to,   100 * p_to,
      ifelse(is.na(tie_rate), "", sprintf("<br>Tie rate: %.2f%%", 100 * tie_rate))
    ),
    smooth = FALSE
  )

visNetwork(nodes2, vn_edges2, width = "100%", height = "800px") %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visPhysics(enabled = FALSE) %>%
  visInteraction(dragView = TRUE, zoomView = TRUE) %>%
  visOptions(
    nodesIdSelection = TRUE,                         # ADDED (search box)
    highlightNearest = list(enabled = TRUE, degree = 1, hover = TRUE)  # ADDED (nice-to-have)
  ) %>%
  visLegend(
    addEdges = data.frame(
      label = levels(pairs$match_bin),
      color = edge_palette
    ),
    addNodes = data.frame(
      label = levels(nodes2$success_bin),
      color = viridisLite::viridis(6)
    ),
    useGroups = FALSE,
    width = 0.25
  )
```

```{r}
library(shiny)
library(bslib)
library(dplyr)
library(tidyr)
library(tibble)
library(rlang)
library(BradleyTerry2)
library(echarts4r)
library(htmlwidgets)

all_categories <- sort(unique(data$category))

ui <- page_fillable(
  theme = bs_theme(bootswatch = "flatly"),
  layout_sidebar(
    sidebar = sidebar(
      selectizeInput("cat", "Filter categories:",
                     choices = all_categories, selected = all_categories,
                     multiple = TRUE, options = list(plugins = list("remove_button"))),
      selectInput("model", "Choose a model:", choices = NULL),
      sliderInput("topn", "Default zoom to Top-N", min = 5, max = 50, value = 10, step = 1)
    ),
    card(
      full_screen = TRUE, collapsible = TRUE, collapsed = FALSE,
      card_header(textOutput("chart_title")),
      card_body(echarts4rOutput("chart", height = "720px"))
    )
  )
)

server <- function(input, output, session){

  filtered_data <- reactive({
    df <- data
    if (!is.null(input$cat) && length(input$cat) > 0) df <- df %>% filter(category %in% input$cat)
    df
  })

  model_order_filtered <- reactive({
    df <- filtered_data()
    appear <- df %>%
      select(model_a, model_b) %>%
      pivot_longer(everything(), values_to = "model") %>%
      count(model, name = "appearances")
    wins <- df %>%
      filter(winner %in% c("model_a","model_b")) %>%
      mutate(winner_model = if_else(winner == "model_a", model_a, model_b)) %>%
      count(winner_model, name = "wins") %>%
      rename(model = winner_model)
    appear %>%
      left_join(wins, by = "model") %>%
      mutate(wins = replace_na(wins, 0),
             success_rate = wins / appearances) %>%
      arrange(desc(success_rate)) %>%
      pull(model)
  })

  observeEvent(model_order_filtered(), {
    choices <- model_order_filtered()
    if (!length(choices)) choices <- sort(unique(c(data$model_a, data$model_b)))
    sel <- if (input$model %in% choices) input$model else choices[1]
    updateSelectInput(session, "model", choices = choices, selected = sel)
  }, ignoreInit = FALSE)

  bt_fit_reactive <- reactive({
    df <- filtered_data() %>%
      filter(winner %in% c("model_a","model_b")) %>%
      transmute(
        player1 = model_a,
        player2 = model_b,
        r1 = as.numeric(winner == "model_a"),
        r2 = as.numeric(winner == "model_b")
      )
    validate(need(nrow(df) > 0, "No decisive matches in the selected filter."))
    players <- sort(unique(c(df$player1, df$player2)))
    df <- df %>% mutate(
      player1 = factor(player1, levels = players),
      player2 = factor(player2, levels = players)
    )
    fit <- tryCatch(BTm(cbind(r1, r2), player1, player2, data = df, id = "player"),
                    error = function(e) NULL)
    validate(need(!is.null(fit), "BT model failed to fit for the selected filter."))
    fit
  })

  abilities_reactive <- reactive({
    fit <- bt_fit_reactive()
    ab <- BTabilities(fit) %>% as.data.frame() %>% tibble::rownames_to_column("model") %>%
      select(model, ability)
    setNames(ab$ability, ab$model)
  })

  pair_ties_reactive <- reactive({
    filtered_data() %>%
      transmute(A = pmin(model_a, model_b),
                B = pmax(model_a, model_b),
                is_tie = as.integer(winner == "tie")) %>%
      filter(A != B) %>%
      summarise(matches = dplyr::n(), ties = sum(is_tie), .by = c(A, B)) %>%
      mutate(tie_rate = ifelse(matches > 0, ties / matches, 0))
  })

  make_vs_table <- function(sel, model_choices, ab_map, pair_ties) {
    opps <- setdiff(model_choices, sel)
    if (!(sel %in% names(ab_map)) || !length(opps))
      return(tibble(model = character(), win_comp = numeric(), tie_half = numeric()))
    tibble(model = opps) %>%
      mutate(
        p_win = ifelse(model %in% names(ab_map),
                       plogis(ab_map[sel] - ab_map[model]), NA_real_),
        A = pmin(sel, model), B = pmax(sel, model)
      ) %>%
      left_join(pair_ties, by = c("A","B")) %>%
      mutate(tie_rate = replace_na(tie_rate, 0),
             p_win = replace_na(p_win, 0)) %>%
      arrange(desc(p_win)) %>%
      transmute(model, win_comp = round(p_win, 3), tie_half = round(0.5 * tie_rate, 3))
  }

  output$chart_title <- renderText({
    cats <- if (is.null(input$cat) || !length(input$cat)) "All categories" else paste(input$cat, collapse = ", ")
    paste0("Predicted vs Others — ", input$model, " (BT win + 0.5× tie)  |  Categories: ", cats)
  })

  output$chart <- renderEcharts4r({
    ab_map   <- abilities_reactive()
    pair_tie <- pair_ties_reactive()
    choices  <- model_order_filtered()
    validate(need(length(choices) > 0, "No models available for the selected filter."))

    df <- make_vs_table(input$model, choices, ab_map, pair_tie) %>%
      mutate(rank = row_number(), label = sprintf("#%d  %s", rank, model))
    validate(need(nrow(df) > 0, "No head-to-head predictions available."))
    top_start <- df$label[1]; top_end <- df$label[min(input$topn, nrow(df))]

    df |>
      e_charts(label) |>
      e_bar(win_comp, name = "P(Win)", stack = "rate",
            label = list(show = TRUE, position = "insideRight", color = "#fff", fontSize = 10)) |>
      e_bar(tie_half, name = "0.5*P(Tie)", stack = "rate",
            label = list(show = TRUE, position = "right", color = "#000", fontSize = 10)) |>
      e_flip_coords() |>
      e_y_axis(inverse = TRUE) |>
      e_legend(orient = "vertical", right = 10, top = 40) |>
      e_title("Predicted Win & Tie Rates Against Other Chatbots") |>
      e_datazoom(y_index = 0, type = "slider", startValue = top_start, endValue = top_end) |>
      e_datazoom(y_index = 0, type = "inside", startValue = top_start, endValue = top_end) |>
      e_x_axis(max = 1,
               axisLabel = list(formatter = htmlwidgets::JS("function(x){return (x*100).toFixed(0)+'%';}"))) |>
      e_axis_labels(y = "Opponent (ranked by P(win))", x = "Success Rate") |>
      e_grid(left = 200) |>
      e_tooltip(show = FALSE) |>
      e_theme("infographic") |>
      e_toolbox_feature("saveAsImage")
  })
}

shinyApp(ui, server)

```

## Limitations

```{r}
# Count matches per pair and per model — sparsity:
df2 %>% count(model_a, model_b) %>% arrange(n)  # look for tiny n
model_summary %>% arrange(matches)  # models with very few matches

# votes per session
session_votes <- flat_data %>%
  count(evaluation_session_id)

ggplot(session_votes, aes(n)) +
  geom_histogram(bins = 40) +
  pub_theme() +
  labs(title = "Votes per Session",
       x = "Votes in Session", y = "Count of Sessions")

# Distribution of judges / sessions: are a small set of users contributing many votes? If you have evaluation_session_id or timestamp, check duplicates:
df2 %>% count(evaluation_session_id) %>% top_n(20, n)

# Order effect (evaluation_order): does the first-shown model get more votes?
df2 %>% group_by(evaluation_order) %>%
  summarise(win_rate_first = mean(if_else(evaluation_order == "A", a_win, b_win)))

# Time trends: models have versions; check timestamps — a model may improve over time.
library(lubridate)
df2 %>% mutate(ts = ymd_hms(timestamp)) %>% arrange(ts) %>% 
  NULL

model_ts <- flat_data %>%
  filter(model_a == "Claude 3.5 Sonnet") %>%   # example — you can parametrize this
  mutate(date = as.Date(timestamp)) %>%
  group_by(date) %>%
  summarise(success = mean(winner == "model_a" | winner == "tie"*0.5))

ggplot(model_ts, aes(date, success)) +
  geom_line() +
  geom_smooth(se = FALSE) +
  pub_theme() +
  labs(title = "Time Series Win Rate for Claude 3.5 Sonnet",
       x = "Date", y = "Success Rate")
```

\[insert discussion\]

Short list for your write-up (explain each briefly): Selection bias of prompts: The set of prompts users choose to run is not random. Users may cherry-pick prompts that favor particular models (e.g., models known to be strong at coding get coding prompts). Non-random assignment of pairings: Some models may be pitted against weaker opponents more often. Non-independent judges / vote stuffing: A small set of users could generate many matches and bias results (or coordinated voting). Presentation / ordering effects: The interface might show Model A on the left or highlight one model first; humans have position/anchoring biases. Anonymity revealed post-vote: If users learn model identity after voting, future votes may be biased by familiarity or brand. Ties/Both Bad handling: How ties are scored affects rankings (half-win vs. remove). Model versions change over time: The dataset aggregates across time; newer model versions likely better — mixing versions mixes abilities. Different prompt difficulty across categories: A model may appear strong simply because it faced easier prompts in a category. Judge expertise and calibration: Users differ in domain expertise; uncalibrated judgments (one user prefers different style) add measurement noise. How to game it: Create many accounts / sessions and push desired model to win. Submit prompts that expose known weaknesses in a competitor. Coordinate to flood-match a model against a weak competitor to inflate win-rate. Exploit UI order bias (always vote for left-side model).

```{r}
library(shiny)
library(bslib)
library(dplyr)
library(tidyr)
library(tibble)
library(purrr)
library(scales)
library(echarts4r)
library(htmlwidgets)

stopifnot("category" %in% names(data))

# ---------- helpers ----------
# Expand each match into (model, category, win/tie) rows for both sides
make_df2 <- function(df) {
  df %>%
    mutate(
      a_win = as.integer(winner == "model_a"),
      b_win = as.integer(winner == "model_b"),
      a_tie = as.integer(winner == "tie"),
      b_tie = a_tie
    ) %>%
    pivot_longer(c(model_a, model_b), names_to = "side", values_to = "model") %>%
    mutate(
      win = if_else(side == "model_a", a_win, b_win),
      tie = if_else(side == "model_a", a_tie, b_tie)
    )
}

# Wilson CI for a proportion (works for success rate or win rate)
wilson_ci <- function(successes, n, z = 1.96) {
  ifelse(n > 0, {
    phat <- successes / n
    denom <- 1 + (z^2)/n
    center <- (phat + (z^2)/(2*n)) / denom
    half   <- (z * sqrt((phat*(1-phat) + (z^2)/(4*n))/n)) / denom
    tibble(lower = pmax(0, center - half), upper = pmin(1, center + half))
  }, tibble(lower = NA_real_, upper = NA_real_))
}

all_models <- sort(unique(c(data$model_a, data$model_b)))
all_categories <- sort(unique(data$category))

# ---------- UI ----------
ui <- page_fillable(
  theme = bs_theme(bootswatch = "flatly"),
  layout_sidebar(
    sidebar = sidebar(
      selectInput("model", "Choose a model:", choices = all_models, selected = all_models[1]),
      selectizeInput("cats", "Categories:", choices = all_categories, selected = all_categories,
                     multiple = TRUE, options = list(plugins = list("remove_button"))),
      sliderInput("min_n", "Min matches per category", min = 1, max = 200, value = 25, step = 1),
      checkboxInput("use_success", "Use Success Rate (Win + 0.5×Tie)", value = TRUE),
      checkboxInput("show_ci", "Show 95% CIs in table", value = TRUE)
    ),
    card(
      full_screen = TRUE, collapsible = TRUE, collapsed = FALSE,
      card_header(textOutput("title_bar")),
      card_body(echarts4rOutput("bar", height = "520px"))
    ),
    card(
      full_screen = FALSE, collapsible = TRUE, collapsed = FALSE,
      card_header("Strengths & Weaknesses (Radar; normalized within selected categories)"),
      card_body(echarts4rOutput("radar", height = "420px"))
    ),
    card(
      collapsible = TRUE, collapsed = TRUE,
      card_header("Details by Category"),
      card_body(tableOutput("detail_table"))
    )
  )
)

# ---------- SERVER ----------
server <- function(input, output, session){

  # Filter to chosen categories
  filtered <- reactive({
    df <- data %>% filter(category %in% input$cats)
    validate(need(nrow(df) > 0, "No rows in selected categories."))
    df
  })

  # Long form per (model, category)
  df2 <- reactive({ make_df2(filtered()) })

  # Per-(model, category) summary
  model_cat_summary <- reactive({
    df2() %>%
      group_by(model, category) %>%
      summarise(
        wins = sum(win),
        ties = sum(tie),
        matches = n(),
        win_rate = wins / matches,
        tie_rate = ties / matches,
        success_rate = (wins + 0.5 * ties) / matches,
        .groups = "drop"
      )
  })

  # Selected model profile + category benchmark (weighted by per-model matches)
  sel_profile <- reactive({
    metric <- if (input$use_success) "success_rate" else "win_rate"
    mcs <- model_cat_summary()

    me <- mcs %>%
      filter(model == input$model) %>%
      rename(model_matches = matches)

    bench <- mcs %>%
      group_by(category) %>%
      summarise(
        bench = weighted.mean(.data[[metric]], w = .data$matches),
        total_matches = sum(.data$matches),
        .groups = "drop"
      )

    me %>%
      inner_join(bench, by = "category") %>%
      mutate(
        metric = .data[[metric]],
        lift = metric - bench
      ) %>%
      filter(model_matches >= input$min_n) %>%
      arrange(desc(metric))
  })

  # Title
  output$title_bar <- renderText({
    paste0("Category Profile for ", input$model, " — ",
           if (input$use_success) "Success Rate (Win + 0.5×Tie)" else "Win Rate",
           "  |  Min matches: ", input$min_n)
  })

  # BAR: model vs category average (with lift tooltip)
  output$bar <- renderEcharts4r({
    prof <- sel_profile()
  
    # Drop rows with missing metric/bench; require finite values
    prof <- prof %>%
      filter(is.finite(metric), is.finite(bench)) %>%
      # still guard against empty after filtering
      mutate(
        metric = ifelse(is.na(metric), 0, metric),
        bench  = ifelse(is.na(bench),  0, bench)
      )
  
    validate(need(nrow(prof) > 0, "No categories meet the minimum matches threshold."))
  
    chart_df <- prof %>%
      arrange(desc(metric)) %>%
      mutate(
        cat_lab = sprintf("%s  (n=%d)", category, model_matches),
        my_val  = round(metric, 6),   # keep numeric; rounding later in JS
        bench_r = round(bench,  6)
      )
  
    chart_df |>
      e_charts(cat_lab) |>
      e_bar(
        my_val, name = input$model,
        label = list(
          show = TRUE, position = "insideRight",
          color = "#fff", fontSize = 10,
          formatter = htmlwidgets::JS(
            "function(p){
               if(p.value == null || !isFinite(p.value)) return '';
               return (p.value*100).toFixed(0) + '%';
             }")
        )
      ) |>
      e_line(
        bench_r, name = "Category Avg", y_index = 0,
        symbol = "circle", symbolSize = 6
      ) |>
      e_tooltip(
        trigger = "axis",
        formatter = htmlwidgets::JS(
          "function (params) {
             const bar  = params.find(p => p.seriesName !== 'Category Avg');
             const line = params.find(p => p.seriesName === 'Category Avg');
             const vBar  = (bar  && isFinite(bar.value))  ? bar.value  : 0;
             const vLine = (line && isFinite(line.value)) ? line.value : 0;
             const lift  = (vBar - vLine)*100;
             return params[0].axisValue + '<br/>' +
                    bar.seriesName + ': ' + (vBar*100).toFixed(1) + '%<br/>' +
                    'Category Avg: ' + (vLine*100).toFixed(1) + '%<br/>' +
                    'Lift: ' + lift.toFixed(1) + ' pts';
           }")
      ) |>
      e_flip_coords() |>
      e_y_axis(inverse = TRUE) |>
      e_title("Model vs Category Average") |>
      e_x_axis(
        max = 1,
        axisLabel = list(formatter = htmlwidgets::JS(
          "function(x){ return (x*100).toFixed(0)+'%'; }"))
      ) |>
      e_axis_labels(
        y = "Category (n = matches)",
        x = if (input$use_success) "Success Rate" else "Win Rate"
      ) |>
      e_grid(left = 170) |>
      e_legend(right = 10, top = 30) |>
      e_datazoom(y_index = 0, type = "slider") |>
      e_theme("infographic") |>
      e_toolbox_feature("saveAsImage")
  })


  # RADAR: normalized strengths (0–1 within selected categories)
  output$radar <- renderEcharts4r({
    mcs <- model_cat_summary() %>% filter(matches >= input$min_n)
    metric <- if (input$use_success) "success_rate" else "win_rate"
  
    # Compute min/max per category for normalization
    cat_minmax <- mcs %>%
      group_by(category) %>%
      summarise(
        mn = suppressWarnings(min(.data[[metric]], na.rm = TRUE)),
        mx = suppressWarnings(max(.data[[metric]], na.rm = TRUE)),
        .groups = "drop"
      )
  
    me <- mcs %>%
      filter(model == input$model) %>%
      inner_join(cat_minmax, by = "category") %>%
      mutate(
        # Normalize; if mx==mn or NA, set to 0.5 neutral
        val = dplyr::case_when(
          is.finite(mx) & is.finite(mn) & mx > mn ~ ( .data[[metric]] - mn ) / (mx - mn),
          TRUE ~ 0.5
        )
      ) %>%
      arrange(category)
  
    # Keep only categories with finite values
    me <- me %>% filter(is.finite(val))
  
    validate(need(nrow(me) >= 3, "Not enough categories with sufficient matches for radar."))
  
    # Wide one-row frame; replace any remaining NA with 0.5
    df_wide <- me %>%
      select(category, val) %>%
      tidyr::pivot_wider(names_from = category, values_from = val) %>%
      mutate(across(everything(), ~ ifelse(is.na(.x), 0.5, .x))) %>%
      mutate(model = input$model)
  
    # Category columns for the series
    cat_cols <- setdiff(names(df_wide), "model")
    validate(need(length(cat_cols) >= 3, "Radar needs ≥ 3 categories."))
  
    # Indicator list-of-lists to avoid jsonlite named-vector warnings
    indicator_list <- lapply(cat_cols, function(nm) list(name = nm, max = 1))
  
    df_wide |>
      e_charts(model) |>
      e_radar(!!!rlang::syms(cat_cols), name = input$model) |>
      e_radar_opts(indicator = indicator_list) |>
      e_title("Normalized Strengths by Category (0–1 within chosen categories)") |>
      e_tooltip(trigger = "item") |>
      e_theme("infographic")
  })


  # DETAILS table (with optional 95% Wilson CI)
  output$detail_table <- renderTable({
    prof <- sel_profile()
    metric <- if (input$use_success) "success_rate" else "win_rate"

    if (input$use_success) {
      ci <- wilson_ci(successes = prof$wins + 0.5 * prof$ties, n = prof$model_matches)
    } else {
      ci <- wilson_ci(successes = prof$wins, n = prof$model_matches)
    }

    tibble(
      Category = prof$category,
      `Matches (model)` = prof$model_matches,
      Wins = prof$wins,
      Ties = prof$ties,
      Rate = scales::percent(prof$metric, accuracy = 0.1),
      `95% CI` = if (input$show_ci)
        paste0(scales::percent(ci$lower, 0.1), " – ", scales::percent(ci$upper, 0.1))
        else NA_character_,
      `Category Avg` = scales::percent(prof$bench, 0.1),
      `Lift (pts)` = round(100 * prof$lift, 1),
      `Matches (category total)` = prof$total_matches
    )
  },
  striped = TRUE,
  bordered = FALSE,
  spacing = "m",
  align = "lrrrccrrr"  # 9 columns → one alignment char per column
  )
}

shinyApp(ui, server)



```
