---
title: "Chatbot Ratings"
output: html_document
runtime: shiny
editor: visual
---

## Introduction

Chatbot Arena (lmarena.ai) is "an open-source platform for evaluating AI through human preference, developed by researchers at UC Berkeley [SkyLab](https://sky.cs.berkeley.edu/) and [LMSYS](https://lmsys.org/)." On the website <https://lmarena.ai/>, users can converse with two LLMs on any topic they choose. Users can then vote on which response they prefer (Model A or Model B) or can choose "Tie" or "Both bad" if they don't have a preference for either model. The LLMs are initially anonymous, but are revealed to the user after they vote.

## Data

The developers of this site provide data [here](https://huggingface.co/datasets/lmarena-ai/arena-human-preference-140k) on 140K conversations with pairwise human preferences. You can find this data in [this shared folder](https://yaleedu-my.sharepoint.com/:f:/g/personal/brian_macdonald_yale_edu/Eqo7cptmHAZAjNMv2F1dCj4BhZzs17e4hkfxbH8tMxLnwQ?e=Oo1DNN). There are 7 `.parquet` files. There is a README file in the GitHub repo that gives details about the data.

Each row is a matchup between two chat bots (e.g. Gemini 2.5 Pro vs Claude Sonnet 4), and the columns indicate a conversation ID, the names of the two models, who the user selected as the winner of the matchup, a common query that the two models were given, the category of the prompt (creative writing, math, code, etc.) and other information. See the README file for more details.

## Question

Which chatbot(s) appears to perform best in the eyes of the judges/users? In a matchup between any two models Model A and Model B, what are the outcome probabilities? How do these change if you focus on specific categories (e.g. writing, math, code, etc.) of prompts? Can you think of any reason the data may not be reliable, or any way this rating system could be "gamed"?

## Submission

Submit a PDF report with your code and analysis to Gradescope.

## Setup

```{r, eval=FALSE}
# read in data and bind rows
datalist <- vector(mode = "list", length = 7)

for(i in 0:6){
  data <- read_parquet(paste0("data/train-0000", i, "-of-00007.parquet")) %>% 
    as.data.frame() %>% 
    select(c(1:6, 11:14))
  datalist[[i+1]] <- data
  print(i)
}

df <- bind_rows(datalist)

flattened_data <- df %>% 
  unpack(cols = category_tag) %>%
  unpack(cols = c(creative_writing_v0.1, criteria_v0.1, if_v0.1, math_v0.1), 
         names_sep = "_")

saveRDS(flattened_data, file = "data.rds")
```

```{r}
# load packages
library(arrow)
library(dplyr)
library(tidyr)
library(stringr)
library(purrr)
library(ggplot2)
library(pubtheme)
library(tidyverse)
library(igraph)
library(ggraph)
library(tidygraph)
library(visNetwork)
library(scales)
library(viridisLite)
library(echarts4r)
library(BradleyTerry2)

# load data
data <- readRDS("data.rds")
head(data)
colnames(data)
```

## Best Chatbot

```{r}
df2 <- data %>%
  mutate(
    a_win = case_when(winner == "model_a" ~ 1, TRUE ~ 0),
    b_win = case_when(winner == "model_b" ~ 1, TRUE ~ 0),
    a_tie = case_when(winner == "tie" ~ 1, TRUE ~ 0),
    b_tie = a_tie
  ) %>%
  pivot_longer(
    cols = c(model_a, model_b),
    names_to = "side",
    values_to = "model"
  ) %>%
  mutate(
    win = if_else(side == "model_a", a_win, b_win),
    tie = if_else(side == "model_a", a_tie, b_tie)
  )

model_summary <- df2 %>%
  group_by(model) %>%
  summarise(
    wins = sum(win),
    ties = sum(tie),
    matches = n(), 
    win_rate = wins / matches,
    tie_rate = ties / matches,
    success_rate = (wins + 0.5 * ties) / matches,
    .groups = "drop"
  ) %>%
  arrange(desc(success_rate))

model_summary

# chart
leaderboard <- model_summary %>%
  arrange(desc(success_rate)) %>%
  mutate(
    rank     = row_number(),
    label    = sprintf("#%d  %s", rank, model),
    win_comp = round(win_rate,3),
    tie_half = round(0.5 * tie_rate,3)
  )

top10_start <- leaderboard$label[1]
top10_end   <- leaderboard$label[min(10, nrow(leaderboard))]

leaderboard |>
  e_charts(label) |>
  e_bar(
    win_comp, name = "Win Rate", stack = "rate",
    label = list(
      show = TRUE, position = "insideRight",
      color = "#fff", fontSize = 10)
  ) |>
  e_bar(
    tie_half, name = "0.5*Tie Rate", stack = "rate",
    label = list(
      show = TRUE, position = "insideRight",
      color = "#fff", fontSize = 10)
  ) |>
  e_flip_coords() |>
  e_y_axis(inverse = TRUE) |>
  e_legend(orient = "vertical", right = 10, top = 40) |>
  e_title("Chatbots Ranked by Success Rate (Win + 0.5*Tie)") |>
  e_datazoom(y_index = 0, type = "slider",
             startValue = top10_start, endValue = top10_end) |>
  e_datazoom(y_index = 0, type = "inside",
             startValue = top10_start, endValue = top10_end) |>
  e_x_axis(max = 1,
           axisLabel = list(formatter = htmlwidgets::JS("function(x){return (x*100).toFixed(0)+'%';}"))) |>
  e_axis_labels(y = "Model (ranked)", x = "Success Rate") |>
  e_grid(left = 170) |>
  e_tooltip(show = FALSE) |>
  e_theme("infographic") |>
  e_toolbox_feature("saveAsImage")
```

## Match-Up Outcome Probabilities

```{r}
# Filter out ties for standard BT model (ties need a separate extension)
bt_data <- data %>%
  filter(winner %in% c("model_a", "model_b")) %>%
  transmute(
    player1 = model_a,
    player2 = model_b,
    result1 = as.numeric(winner == "model_a"),
    result2 = as.numeric(winner == "model_b")
  )
# ensure same factor levels for both sides
players <- sort(unique(c(bt_data$player1, bt_data$player2)))
bt_data <- bt_data %>%
  mutate(
    player1 = factor(player1, levels = players),
    player2 = factor(player2, levels = players)
  )

bt_model <- BTm(
  cbind(result1, result2), player1, player2,
  data = bt_data, id = "player"
)

summary(bt_model)

abilities <- BTabilities(bt_model) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("model") %>%
  arrange(desc(ability)) %>%
  mutate(
    strength = exp(ability),        # convert to positive “skill” scale
    normalized = strength / sum(strength)
  )

head(abilities, 10)

predict_bt <- function(modelA, modelB, abilities_df = abilities) {
  a <- abilities_df$ability[abilities_df$model == modelA]
  b <- abilities_df$ability[abilities_df$model == modelB]
  pA <- exp(a) / (exp(a) + exp(b))
  tibble(modelA, modelB, P_A_wins = pA, P_B_wins = 1 - pA)
}
```

```{r}
# 1) order by success_rate (highest → lowest)
model_order <- model_summary %>% arrange(desc(success_rate)) %>% pull(model)

# keep only models present in the BT abilities table
model_order <- intersect(model_order, abilities$model)

# 2) ability lookup
ab_map <- setNames(abilities$ability, abilities$model)

# 3) full grid of pairs with BT prob P(row beats col)
grid <- expand.grid(
  row_model = model_order,
  col_model = model_order,
  stringsAsFactors = FALSE
) %>%
  mutate(
    prob = ifelse(
      row_model == col_model, NA_real_,
      plogis(ab_map[row_model] - ab_map[col_model])  # exp(a)/(exp(a)+exp(b))
    )
  )

# factors to get the desired ordering in the plot:
grid$row_model <- factor(grid$row_model, levels = rev(model_order))  # rows: top = best
grid$col_model <- factor(grid$col_model, levels = model_order)       # cols: left = best

# 4) static heatmap
p <- ggplot(grid, aes(x = col_model, y = row_model, fill = prob)) +
  geom_tile(color = "white", linewidth = 0.2, na.rm = FALSE) +
  scale_fill_viridis_c(
    option = "D", limits = c(0, 1),
    labels = percent_format(accuracy = 1),
    name = "P(Row beats Col)"
  ) +
  coord_equal() +
  labs(
    title = "Head-to-Head Win Probabilities (Bradley–Terry)",
    x = "Column model", y = "Row model"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(size = 4, angle = 45, hjust = 1, vjust = 1),
    axis.text.y = element_text(size = 4)
  )

print(p)
```

```{r}
# edges and nodes
pairs <- data %>% 
  count(model_a, model_b, name = "matches") %>%
  mutate(
    edge_width = log1p(matches),   # damp outliers for thickness
    match_bin = cut(                # edge COLOR uses raw-match bins (for legend)
      matches,
      breaks = c(1,5,10,25,50,100,Inf),
      labels = c("1–5","6–10","11–25","26–50","51–100","100+"),
      right = TRUE, include.lowest = TRUE
    )
  )

nodes <- model_summary %>% 
  select(model, success_rate)

g <- graph_from_data_frame(d = pairs, vertices = nodes, directed = FALSE)

# ---- ADDED: total matches per model (sum of incident edge matches) ----------
totals <- pairs %>%
  select(model_a, model_b, matches) %>%
  pivot_longer(c(model_a, model_b), names_to = "side", values_to = "model") %>%
  group_by(model) %>%
  summarise(total_matches = sum(matches, na.rm = TRUE), .groups = "drop")

# success rate bins for legend + node popups
nodes2 <- model_summary %>%
  select(model, success_rate) %>%
  rename(id = model) %>%
  left_join(totals, by = c("id" = "model")) %>%     # ADDED
  mutate(
    success_bin = cut(
      success_rate,
      breaks = seq(min(success_rate, na.rm = TRUE),  # ADDED: na.rm for safety
                   max(success_rate, na.rm = TRUE),
                   length.out = 7),
      include.lowest = TRUE
    ),
    color = viridisLite::viridis(6)[as.integer(success_bin)],
    label = id,
    font.color = color,
    # ADDED: tooltip content
    title = paste0(
      "<b>", id, "</b><br>",
      "Success rate: ", percent(success_rate, accuracy = 0.1), "<br>",
      "Total matches: ", ifelse(is.na(total_matches), 0, total_matches)
    )
  )

# --- NEW: empirical head-to-head outcome probabilities per unordered pair ----
edge_probs <- data %>%
  transmute(
    A = pmin(model_a, model_b),
    B = pmax(model_a, model_b),
    res = case_when(
      winner == "tie" ~ "tie",
      winner == "model_a" & model_a <= model_b ~ "A_win",
      winner == "model_b" & model_a  > model_b ~ "A_win",
      winner == "model_b" & model_a <= model_b ~ "B_win",
      winner == "model_a" & model_a  > model_b ~ "B_win",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(res), A != B) %>%
  count(A, B, res) %>%
  tidyr::pivot_wider(
    names_from = res, values_from = n, values_fill = 0
  ) %>%
  mutate(
    total = A_win + B_win + tie,
    pA = ifelse(total > 0, A_win / total, NA_real_),  # P(A beats B)
    pB = ifelse(total > 0, B_win / total, NA_real_),  # P(B beats A)
    pT = ifelse(total > 0, tie   / total, NA_real_)   # tie rate
  )

# --- NEW: join probabilities onto your existing pairs in a direction-aware way
pairs_with_probs <- pairs %>%
  mutate(A = pmin(model_a, model_b),
         B = pmax(model_a, model_b)) %>%
  left_join(edge_probs, by = c("A","B")) %>%
  mutate(
    from = model_a, to = model_b,
    p_from = if_else(model_a <= model_b, pA, pB),
    p_to   = if_else(model_a <= model_b, pB, pA),
    tie_rate = pT
  )

# darker greys, first one darker manually (keep your palette)
edge_palette <- RColorBrewer::brewer.pal(6, "Greys")
edge_palette[1] <- "#6f6f6f"

vn_edges2 <- pairs_with_probs %>%
  transmute(
    from, to,
    width = pmax(1, round(log1p(matches))),
    color = edge_palette[as.integer(match_bin)],
    # Edge label now includes matchups + probabilities (rounded to 2dp)
    title = sprintf(
      "<b>%s ↔ %s</b><br>Matches: %d<br>%s win prob: %.2f%%<br>%s win prob: %.2f%%%s",
      from, to, matches,
      from, 100 * p_from,
      to,   100 * p_to,
      ifelse(is.na(tie_rate), "", sprintf("<br>Tie rate: %.2f%%", 100 * tie_rate))
    ),
    smooth = FALSE
  )

visNetwork(nodes2, vn_edges2, width = "100%", height = "800px") %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visPhysics(enabled = FALSE) %>%
  visInteraction(dragView = TRUE, zoomView = TRUE) %>%
  visOptions(
    nodesIdSelection = TRUE,                         # ADDED (search box)
    highlightNearest = list(enabled = TRUE, degree = 1, hover = TRUE)  # ADDED (nice-to-have)
  ) %>%
  visLegend(
    addEdges = data.frame(
      label = levels(pairs$match_bin),
      color = edge_palette
    ),
    addNodes = data.frame(
      label = levels(nodes2$success_bin),
      color = viridisLite::viridis(6)
    ),
    useGroups = FALSE,
    width = 0.25
  )
```

```{r}
library(shiny)
library(bslib)
# 1) Order models by success rate (for dropdown & consistent views)
model_order <- model_summary %>% arrange(desc(success_rate)) %>% pull(model)

# 2) Bradley–Terry abilities (ties removed for BT fit)
bt_data <- data %>%
  filter(winner %in% c("model_a","model_b")) %>%
  transmute(
    player1 = model_a,
    player2 = model_b,
    r1 = as.numeric(winner == "model_a"),
    r2 = as.numeric(winner == "model_b")
  )

players <- sort(unique(c(bt_data$player1, bt_data$player2)))
bt_data <- bt_data %>%
  mutate(
    player1 = factor(player1, levels = players),
    player2 = factor(player2, levels = players)
  )

bt_model <- BTm(cbind(r1, r2), player1, player2, data = bt_data, id = "player")
abilities <- BTabilities(bt_model) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("model") %>%
  select(model, ability)

ab_map <- setNames(abilities$ability, abilities$model)

# 3) Empirical tie rate per unordered pair (used for 0.5 * tie stack)
pair_ties <- data %>%
  transmute(A = pmin(model_a, model_b),
            B = pmax(model_a, model_b),
            is_tie = as.integer(winner == "tie")) %>%
  filter(A != B) %>%
  summarise(matches = dplyr::n(), ties = sum(is_tie), .by = c(A, B)) %>%
  mutate(tie_rate = ifelse(matches > 0, ties / matches, 0))

# Helper: table for a selected model (predicted BT win vs each opponent)
make_vs_table <- function(sel) {
  opps <- setdiff(model_order, sel)
  tibble(model = opps) %>%
    mutate(
      p_win = plogis(ab_map[sel] - ab_map[model]),    # BT predicted P(sel beats model)
      A = pmin(sel, model),
      B = pmax(sel, model)
    ) %>%
    left_join(pair_ties, by = c("A","B")) %>%
    mutate(tie_rate = ifelse(is.na(tie_rate), 0, tie_rate)) %>%
    arrange(desc(p_win)) %>%
    transmute(
      model,                                  # opponent
      win_comp = round(p_win, 3),             # keep your rounding
      tie_half = round(0.5 * tie_rate, 3)     # match your stack definition
    )
}

# ================================ SHINY APP ===================================
ui <- page_fillable(
  theme = bs_theme(bootswatch = "flatly"),
  layout_sidebar(
    sidebar = sidebar(
      selectInput("model", "Choose a model:", choices = model_order, selected = model_order[1]),
      sliderInput("topn", "Default zoom to Top-N", min = 5, max = 50, value = 10, step = 1)
    ),
    card(
      full_screen = TRUE, collapsible = TRUE, collapsed = FALSE,
      card_header(textOutput("chart_title")),
      card_body(echarts4rOutput("chart", height = "720px"))
    )
  )
)

server <- function(input, output, session){

  output$chart_title <- renderText({
    paste0("Predicted vs Others — ", input$model, " (BT win + 0.5× tie)")
  })

  output$chart <- renderEcharts4r({
    df <- make_vs_table(input$model) %>%
      mutate(
        rank = row_number(),
        label = sprintf("#%d  %s", rank, model)  # << your label field
      )

    top10_start <- df$label[1]
    top10_end   <- df$label[min(input$topn, nrow(df))]

    # ---- your chart, unchanged except data source/labels ---------------------
    df |>
      e_charts(label) |>
      e_bar(
        win_comp, name = "P(Win)", stack = "rate",
        label = list(show = TRUE, position = "insideRight",
                     color = "#fff", fontSize = 10)
      ) |>
      e_bar(
        tie_half, name = "0.5*P(Tie)", stack = "rate",
        label = list(show = TRUE, position = "right",
                     color = "#000", fontSize = 10)
      ) |>
      e_flip_coords() |>
      e_y_axis(inverse = TRUE) |>
      e_legend(orient = "vertical", right = 10, top = 40) |>
      e_title("Predicted Win & Tie Rates Against Other Chatbots") |>
      e_datazoom(y_index = 0, type = "slider",
                 startValue = top10_start, endValue = top10_end) |>
      e_datazoom(y_index = 0, type = "inside",
                 startValue = top10_start, endValue = top10_end) |>
      e_x_axis(max = 1,
               axisLabel = list(formatter = htmlwidgets::JS("function(x){return (x*100).toFixed(0)+'%';}"))) |>
      e_axis_labels(y = "Opponent (ranked by P(win))", x = "Success Rate") |>
      e_grid(left = 200) |>
      e_tooltip(show = FALSE) |>
      e_theme("infographic") |>
      e_toolbox_feature("saveAsImage")
  })
}

shinyApp(ui, server)
```

## Limitations

```{r}
# Count matches per pair and per model — sparsity:
df2 %>% count(model_a, model_b) %>% arrange(n)  # look for tiny n
model_summary %>% arrange(matches)  # models with very few matches

# votes per session
session_votes <- flat_data %>%
  count(evaluation_session_id)

ggplot(session_votes, aes(n)) +
  geom_histogram(bins = 40) +
  pub_theme() +
  labs(title = "Votes per Session",
       x = "Votes in Session", y = "Count of Sessions")

# Distribution of judges / sessions: are a small set of users contributing many votes? If you have evaluation_session_id or timestamp, check duplicates:
df2 %>% count(evaluation_session_id) %>% top_n(20, n)

# Order effect (evaluation_order): does the first-shown model get more votes?
df2 %>% group_by(evaluation_order) %>%
  summarise(win_rate_first = mean(if_else(evaluation_order == "A", a_win, b_win)))

# Time trends: models have versions; check timestamps — a model may improve over time.
library(lubridate)
df2 %>% mutate(ts = ymd_hms(timestamp)) %>% arrange(ts) %>% 
  NULL

model_ts <- flat_data %>%
  filter(model_a == "Claude 3.5 Sonnet") %>%   # example — you can parametrize this
  mutate(date = as.Date(timestamp)) %>%
  group_by(date) %>%
  summarise(success = mean(winner == "model_a" | winner == "tie"*0.5))

ggplot(model_ts, aes(date, success)) +
  geom_line() +
  geom_smooth(se = FALSE) +
  pub_theme() +
  labs(title = "Time Series Win Rate for Claude 3.5 Sonnet",
       x = "Date", y = "Success Rate")
```

\[insert discussion\]

Short list for your write-up (explain each briefly): Selection bias of prompts: The set of prompts users choose to run is not random. Users may cherry-pick prompts that favor particular models (e.g., models known to be strong at coding get coding prompts). Non-random assignment of pairings: Some models may be pitted against weaker opponents more often. Non-independent judges / vote stuffing: A small set of users could generate many matches and bias results (or coordinated voting). Presentation / ordering effects: The interface might show Model A on the left or highlight one model first; humans have position/anchoring biases. Anonymity revealed post-vote: If users learn model identity after voting, future votes may be biased by familiarity or brand. Ties/Both Bad handling: How ties are scored affects rankings (half-win vs. remove). Model versions change over time: The dataset aggregates across time; newer model versions likely better — mixing versions mixes abilities. Different prompt difficulty across categories: A model may appear strong simply because it faced easier prompts in a category. Judge expertise and calibration: Users differ in domain expertise; uncalibrated judgments (one user prefers different style) add measurement noise. How to game it: Create many accounts / sessions and push desired model to win. Submit prompts that expose known weaknesses in a competitor. Coordinate to flood-match a model against a weak competitor to inflate win-rate. Exploit UI order bias (always vote for left-side model).
