---
title: "Chatbot Ratings"
format: pdf
editor: visual
---

## Introduction

Chatbot Arena (lmarena.ai) is "an open-source platform for evaluating AI through human preference, developed by researchers at UC Berkeley [SkyLab](https://sky.cs.berkeley.edu/) and [LMSYS](https://lmsys.org/)." On the website <https://lmarena.ai/>, users can converse with two LLMs on any topic they choose. Users can then vote on which response they prefer (Model A or Model B) or can choose "Tie" or "Both bad" if they don't have a preference for either model. The LLMs are initially anonymous, but a

re revealed to the user after they vote.

## Data

The developers of this site provide data [here](https://huggingface.co/datasets/lmarena-ai/arena-human-preference-140k) on 140K conversations with pairwise human preferences. You can find this data in [this shared folder](https://yaleedu-my.sharepoint.com/:f:/g/personal/brian_macdonald_yale_edu/Eqo7cptmHAZAjNMv2F1dCj4BhZzs17e4hkfxbH8tMxLnwQ?e=Oo1DNN). There are 7 `.parquet` files. There is a README file in the GitHub repo that gives details about the data.

Each row is a matchup between two chat bots (e.g. Gemini 2.5 Pro vs Claude Sonnet 4), and the columns indicate a conversation ID, the names of the two models, who the user selected as the winner of the matchup, a common query that the two models were given, the category of the prompt (creative writing, math, code, etc.) and other information. See the README file for more details.

## Question

Which chatbot(s) appears to perform best in the eyes of the judges/users? In a matchup between any two models Model A and Model B, what are the outcome probabilities? How do these change if you focus on specific categories (e.g. writing, math, code, etc.) of prompts? Can you think of any reason the data may not be reliable, or any way this rating system could be "gamed"?

## Submission

Submit a PDF report with your code and analysis to Gradescope.

## Setup

```{r}
# read in packages
library(arrow)
library(dplyr)
library(tidyr)
library(stringr)
library(purrr)
library(ggplot2)
library(pubtheme)
library(igraph)
library(ggraph)

# read in data and bind rows
datalist <- vector(mode = "list", length = 7)

for(i in 0:6){
  data <- read_parquet(paste0("data/train-0000", i, "-of-00007.parquet")) %>% 
    as.data.frame() %>% 
    select(c(1:6, 11:14))
  datalist[[i+1]] <- data
  print(i)
}

df <- bind_rows(datalist)

flattened_data <- df %>% 
  unpack(cols = category_tag) %>%
  unpack(cols = c(creative_writing_v0.1, criteria_v0.1, if_v0.1, math_v0.1), 
         names_sep = "_")

saveRDS(flattened_data, file = "data.rds")
```

```{r}
# data exploration
data <- readRDS("data.rds")
head(data)
colnames(data)

table(data$winner, useNA = "ifany")
```

## Best Chatbot

```{r}
df2 <- data %>%
  mutate(
    a_win = case_when(winner == "model_a" ~ 1, TRUE ~ 0),
    b_win = case_when(winner == "model_b" ~ 1, TRUE ~ 0),
    a_tie = case_when(winner == "tie" ~ 1, TRUE ~ 0),
    b_tie = a_tie
  ) %>%
  pivot_longer(
    cols = c(model_a, model_b),
    names_to = "side",
    values_to = "model"
  ) %>%
  mutate(
    win = if_else(side == "model_a", a_win, b_win),
    tie = if_else(side == "model_a", a_tie, b_tie)
  )

model_summary <- df2 %>%
  group_by(model) %>%
  summarise(
    wins = sum(win),
    ties = sum(tie),
    matches = n(), 
    win_rate = wins / matches,
    tie_rate = ties / matches,
    success_rate = (wins + 0.5 * ties) / matches,
    .groups = "drop"
  ) %>%
  arrange(desc(success_rate))

model_summary

# chart
top10 <- model_summary %>%
  slice_max(success_rate, n = 10) %>%
  arrange(desc(success_rate)) %>%
  mutate(model = reorder(model, success_rate))

ggplot(top10, aes(x = model)) +
  geom_col(aes(y = win_rate, fill = "Wins")) +
  geom_col(aes(y = 0.5 * tie_rate, fill = "0.5 Tie Contribution")) +
  geom_text(aes(y = success_rate, label = round(success_rate,3)), size = 3.5) +
  coord_flip(clip = "off") +
  scale_fill_manual(
    name = "Component",
    values = c("Wins" = "blue", "0.5 Tie Contribution" = "red")
  ) +
  labs(
    title = "Top 10 Chatbots by Success Rate (Win + 0.5*Tie)",
    x = "Model",
    y = "Success Rate"
  ) +
  theme_pub() +
  theme(plot.margin = margin(r = 15))
```

## Match-Up Outcome Probabilities

```{r}
# visualize match-ups
pair_counts <- data %>%
  count(model_a, model_b)

ggplot(pair_counts, aes(n)) +
  geom_histogram(bins = 30) +
  theme_pub() +
  labs(title = "Distribution of Matches per Model Pair",
       x = "Matches per Pair", y = "Count of Pairs")

# 1) Pair match counts
pairs <- data %>%
  count(model_a, model_b, name="matches")

# 2) node success rates
nodes <- model_summary %>%
  select(model, success_rate)

# 3) build graph object
g <- graph_from_data_frame(d = pairs, vertices = nodes, directed = FALSE)

# 4) plot
ggraph(g, layout = "fr") +
  geom_edge_link(aes(width = matches), alpha = 0.4) +
  geom_node_point(aes(color = success_rate), size = 5) +
  scale_edge_width(range = c(0.2, 3)) +
  scale_color_gradient(low = "red", high = "blue") +  # red bad → blue good (colorblind safe)
  pub_theme() +
  theme(legend.position = "right") +
  labs(
    title = "Chatbot Arena Network",
    subtitle = "Edge thickness = number of matches. Node color = success rate",
    edge_width = "Matches",
    color = "Success Rate"
  )
```

```{r}
pairwise_empirical <- df2 %>%
  group_by(model_a, model_b) %>%
  summarise(
    matches = n(),
    a_wins = sum(a_win),
    b_wins = sum(b_win),
    a_win_rate = a_wins / matches,
    .groups = "drop"
  )

# lookup function:
get_empirical <- function(x, y) {
  pairwise_empirical %>%
    filter(model_a == x, model_b == y) %>%
    select(model_a, model_b, matches, a_win_rate)
}

# example
get_empirical("Gemini 2.5 Pro", "Claude Sonnet 4")

# install.packages("BradleyTerry2") # if needed
library(BradleyTerry2)

# prepare contingency for BT: counts of A wins, B wins, ties handled as half-wins
bt_df <- df2 %>%
  # ensure model names as factors
  mutate(model_a = as.character(model_a), model_b = as.character(model_b)) %>%
  group_by(model_a, model_b) %>%
  summarise(
    A_wins = sum(a_win),
    B_wins = sum(b_win),
    matches = n(),
    .groups = "drop"
  )

# For BradleyTerry2 we need a "long" dataset of wins per pairing:
# create a "contest" id and use counts as responses
bt_long <- bt_df %>%
  mutate(contest = row_number()) %>%
  select(contest, model_a, model_b, A_wins, B_wins, matches)

# Build BT data: form the wins vector and the player factor matrix
# BradleyTerry2 expects (win1, win2) as response, and a player-design matrix
results_list <- with(bt_long, {
  # each row is a contest: cbind(A_wins, B_wins)
  wins <- cbind(A_wins, B_wins)
  players <- data.frame(player1 = model_a, player2 = model_b)
  list(wins = wins, players = players)
})

# Build data for BT
btdata <- BTm(outcome = results_list$wins,
              player1 = results_list$players$player1,
              player2 = results_list$players$player2,
              id = unique(c(results_list$players$player1, results_list$players$player2)))

bt_fit <- btdata
summary(bt_fit)

# extract ability scores (log-ability)
abilities <- BTabilities(bt_fit)
abilities_df <- data.frame(model = rownames(abilities), log_ability = abilities[,1]) %>%
  arrange(desc(log_ability))
abilities_df

# compute pairwise probability function
prob_A_beats_B <- function(modelA, modelB) {
  a <- abilities_df$log_ability[abilities_df$model == modelA]
  b <- abilities_df$log_ability[abilities_df$model == modelB]
  p <- exp(a) / (exp(a) + exp(b))
  p
}

# visualization
# build matrix of pairwise probs from abilities_df
models <- abilities_df$model
prob_mat <- outer(models, models, Vectorize(function(x,y) {
  if(x==y) return(NA_real_)
  prob_A_beats_B(x,y)
}))
colnames(prob_mat) <- rownames(prob_mat) <- models
# convert to long for ggplot
prob_long <- as.data.frame(prob_mat) %>%
  mutate(modelA = rownames(.)) %>%
  pivot_longer(-modelA, names_to = "modelB", values_to = "p")
library(ggplot2)
ggplot(prob_long, aes(x = modelA, y = modelB, fill = p)) +
  geom_tile() + coord_fixed() + theme_minimal() +
  scale_fill_viridis_c(na.value = "white") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

## Match-Up Outcome Probabilities By Category

```{r}
# assume there is a column `creative_writing_v0.1_creative_writing` or `category`
# If you have a single unified category variable, use that; otherwise adapt.

# make a high-level 'category' column. Example heuristic:
df2 <- df2 %>%
  mutate(category = case_when(
    !is.na(creative_writing_v0.1_creative_writing) & creative_writing_v0.1_creative_writing != "" ~ "creative_writing",
    !is.na(math_v0.1_math) & math_v0.1_math != "" ~ "math",
    is_code == TRUE ~ "code",
    TRUE ~ "other"
  ))

# aggregate per category: top models by win-rate
per_cat_summary <- df2 %>%
  filter(winner_norm != "BOTH_BAD") %>%
  group_by(category, model = model_a) %>%
  summarise(wins_from_A = sum(a_win), matches_from_A = n(), .groups="drop_last") %>%
  # need wins contributed by B rows as well; easier to convert to long like before
  ungroup() %>%
  bind_rows(
    df2 %>% mutate(model = model_b, wins = b_win) %>% select(category, model, wins = b_win) %>% group_by(category, model) %>%
      summarise(total_wins = sum(wins), matches = n(), .groups = "drop")
  )

# simpler: short code to compute win-rate per model-category
model_cat <- bind_rows(
  df2 %>% transmute(category, model = model_a, wins = a_win, n = 1),
  df2 %>% transmute(category, model = model_b, wins = b_win, n = 1)
) %>%
  group_by(category, model) %>%
  summarise(total_wins = sum(wins), matches = sum(n), win_rate = total_wins / matches, .groups = "drop") %>%
  arrange(category, desc(win_rate))

model_cat

# visualize
library(boot)

# fn for bootstrap success rate
success_fun <- function(data, idx) {
  d2 <- data[idx,]
  (mean(d2$winner == "model_a") + 0.5 * mean(d2$winner == "tie"))
}

# models per category
cat_results <- flat_data %>%
  mutate(category = language) %>%
  group_by(category, model_a) %>%
  summarise(
    success = success_fun(cur_data_all(), 1:n()),
    ci = list(boot(cur_data_all(), success_fun, R=400))
  ) %>%
  mutate(
    lower = map_dbl(ci, ~ quantile(.x$t, .025)),
    upper = map_dbl(ci, ~ quantile(.x$t, .975))
  ) %>%
  slice_max(success, n = 10) # top 10 per category

ggplot(cat_results, aes(x = reorder(model_a, success), y = success)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0) +
  facet_wrap(~category, scales = "free_y") +
  coord_flip() +
  pub_theme() +
  labs(title = "Top Models per Category with 95% bootstrap CIs",
       x = "Model", y = "Success Rate")
```

## Limitations

```{r}
# Count matches per pair and per model — sparsity:
df2 %>% count(model_a, model_b) %>% arrange(n)  # look for tiny n
model_summary %>% arrange(matches)  # models with very few matches

# votes per session
session_votes <- flat_data %>%
  count(evaluation_session_id)

ggplot(session_votes, aes(n)) +
  geom_histogram(bins = 40) +
  pub_theme() +
  labs(title = "Votes per Session",
       x = "Votes in Session", y = "Count of Sessions")

# Distribution of judges / sessions: are a small set of users contributing many votes? If you have evaluation_session_id or timestamp, check duplicates:
df2 %>% count(evaluation_session_id) %>% top_n(20, n)

# Order effect (evaluation_order): does the first-shown model get more votes?
df2 %>% group_by(evaluation_order) %>%
  summarise(win_rate_first = mean(if_else(evaluation_order == "A", a_win, b_win)))

# Time trends: models have versions; check timestamps — a model may improve over time.
library(lubridate)
df2 %>% mutate(ts = ymd_hms(timestamp)) %>% arrange(ts) %>% 
  NULL

model_ts <- flat_data %>%
  filter(model_a == "Claude 3.5 Sonnet") %>%   # example — you can parametrize this
  mutate(date = as.Date(timestamp)) %>%
  group_by(date) %>%
  summarise(success = mean(winner == "model_a" | winner == "tie"*0.5))

ggplot(model_ts, aes(date, success)) +
  geom_line() +
  geom_smooth(se = FALSE) +
  pub_theme() +
  labs(title = "Time Series Win Rate for Claude 3.5 Sonnet",
       x = "Date", y = "Success Rate")
```

\[insert discussion\]

Short list for your write-up (explain each briefly): Selection bias of prompts: The set of prompts users choose to run is not random. Users may cherry-pick prompts that favor particular models (e.g., models known to be strong at coding get coding prompts). Non-random assignment of pairings: Some models may be pitted against weaker opponents more often. Non-independent judges / vote stuffing: A small set of users could generate many matches and bias results (or coordinated voting). Presentation / ordering effects: The interface might show Model A on the left or highlight one model first; humans have position/anchoring biases. Anonymity revealed post-vote: If users learn model identity after voting, future votes may be biased by familiarity or brand. Ties/Both Bad handling: How ties are scored affects rankings (half-win vs. remove). Model versions change over time: The dataset aggregates across time; newer model versions likely better — mixing versions mixes abilities. Different prompt difficulty across categories: A model may appear strong simply because it faced easier prompts in a category. Judge expertise and calibration: Users differ in domain expertise; uncalibrated judgments (one user prefers different style) add measurement noise. How to game it: Create many accounts / sessions and push desired model to win. Submit prompts that expose known weaknesses in a competitor. Coordinate to flood-match a model against a weak competitor to inflate win-rate. Exploit UI order bias (always vote for left-side model).
