---
title: "ai_analysis_v1"
format: html
editor: visual
---

## Chatbot analysis v1

```{r}
#libraries
library(dplyr)
library(BradleyTerry2)
library(dplyr)
```

Read in the data

```{r}

```

Calculate percentage win rates

```{r}
# count wins per model
model_wins <- d %>%
  filter(winner %in% c("model1", "model2")) %>%
  mutate(winner_model = ifelse(winner == "model1", model1, model2)) %>%
  count(winner_model, name = "wins")

# Count appearances
model_appearances <- d %>%
  select(model1, model2) %>%
  pivot_longer(cols = everything(), values_to = "model") %>%
  count(model, name = "appearances")

# Merge wins and appearances
performance_summary <- model_appearances %>%
  left_join(model_wins, by = c("model" = "winner_model")) %>%
  mutate(wins = replace_na(wins, 0),
         win_rate = wins / appearances) %>%
  arrange(desc(win_rate))

print(performance_summary)
```

Plot win rate of models

```{r}
ggplot(performance_summary, aes(x = reorder(model, win_rate), y = win_rate)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Win Rate of Chatbot Models",
       x = "Model",
       y = "Win Rate (Proportion of Judged Wins)") +
  theme_minimal()
```

Use logistic regression to quantify the performance differences between models while adjusting for factors like opponent or judge. A binary logistic regression was used, with model1 and model 2 as predictors.

```{r}
# Filter to usable rows
df_use <- d %>%
  filter(winner %in% c("model1", "model2"))

# Binary outcome: 1 = model1 won, 0 = model2 won
df_use <- df_use %>%
  mutate(outcome = ifelse(winner == "model1", 1, 0))

# Create indicator variables for model identities
df_use <- df_use %>%
  mutate(
    model1_factor = factor(model1),
    model2_factor = factor(model2)
  )

# Use model1 and model2 as predictors
model <- glm(outcome ~ model1_factor + model2_factor, data = df_use, family = binomial)
summary(model)
```

Bradley-Terry Model (specifically for pairwise comparison data)

```{r}
# Step 1: Prepare pairwise outcome data
df_bt <- d %>%
  filter(winner %in% c("model1", "model2")) %>%
  mutate(
    winner_model = ifelse(winner == "model1", model1, model2),
    loser_model  = ifelse(winner == "model1", model2, model1)
  )

# Step 2: Ensure both winner and loser columns are factors with the same levels
all_models <- union(df_bt$winner_model, df_bt$loser_model)
df_bt <- df_bt %>%
  mutate(
    winner_model = factor(winner_model, levels = all_models),
    loser_model  = factor(loser_model,  levels = all_models)
  )

# Step 3: Create player list for BTm format
players <- data.frame(model = factor(all_models))  # model ID factor

# Step 4: Fit the Bradley-Terry model
bt_model <- BTm(
  outcome = rep(1, nrow(df_bt)),
  player1 = df_bt$winner_model,
  player2 = df_bt$loser_model,
  id = "model",
  data = players
)

summary(bt_model)
```

1.  Subset by fields (math, creative writing, code) - which models perform best for which subsets, which has highest outcome probability for each subset
    1.  build separate bradley-terry models by category
    2.  compare rankings across subsets
2.  Style & length
    1.  How bold affects win rate
    2.  How length affects win rate
    3.  How tone (casual, formal, difficult/simple terminology, etc) affect win rate
3.  Time
    1.  how ratings have changed over time
    2.  identify release inflection points for models
4.  In cases where both models were bad, what were these prompts/the context? where do LLMs struggle still
5.  User behavior
    1.  whether users become harsher/more lenient over a session
    2.  whether they bias model A usually over model B
    3.  when do they not select a clear winner, do some users always choose
    4.  have preferences changed
6.  shiny plot - choose the model and see scores (ex. 90 for code) and strengths and weaknesses
7.  
